---
title: "Common Garden Design"
output: html_notebook
---

## Intro

Determine randomization strategy and plant placing when we have had uneven germination of our original plan.

This is written for UCD Fall 2022 planting.

```{r}
library(tidyverse)
library(googlesheets4)
library(ggforce)

is.even <- function(x) x%%2 == 0
```

## get data on germinated seed

```{r}
# the original reps
seed <- read_sheet("https://docs.google.com/spreadsheets/d/1EaSA2QrlWAjFR35pedcvheQp3X7Agsla6pZtE2C1QJ4/edit#gid=1798047255", 
                   sheet = "germination",
                   na=c("NA", "")
) %>% 
  select(parent.pop, mf, rep, `germinated?`)
seed
```


```{r}
# extra seed
extra <- read_sheet("https://docs.google.com/spreadsheets/d/1EaSA2QrlWAjFR35pedcvheQp3X7Agsla6pZtE2C1QJ4/edit#gid=1798047255", 
                   sheet = "extra mf reps",
                   na=c("NA", "")
) %>%
  filter(`germinated?` == 1) # we are not going to use ungerminated extra seed.
extra
```

```{r}
seed.combined <- bind_rows(seed, extra)

germinated <- seed.combined %>%
  filter(`germinated?`==1)

notgerminated <- seed.combined %>%
  filter(`germinated?` == 0)
```

## assign blocks to germinated

Can plant 160  plants per bed
```{r}
nrow(germinated) 
nrow(germinated) / 160
```

get a histogram of rep count

```{r}
germinated %>%
  group_by(parent.pop, mf) %>%
  summarize(n=n()) %>%
  ggplot(aes(x=n)) +
  geom_histogram(binwidth = 1)
```

```{r}
germinated %>%
  group_by(parent.pop) %>%
  summarize(n=n()) %>%
  arrange(n)
```

A bit challenging because for pops with few reps total we should be careful to distribute the pops across the blocks.

I think one way this can work is to arrange by pops and mfs and then assign a randomized set of block labels to it.

```{r}
# Remember, Bed A is buffer

germbeds <- c("D", "F", "H", "J", "L")
notgermbeds <- LETTERS[2:14] %>% setdiff(germbeds)

germblocks <- str_c(rep(germbeds,2), c("1", "2")) %>% sort

germreps <- ceiling(nrow(germinated) / 10 ) # number of times we have to assign blocks

# set.seed(777)
# germinated <- germinated %>%
#   arrange(parent.pop, mf) %>%
#   mutate(block= {map(1:germreps, ~ sample(germblocks)) %>%
#       unlist() %>% magrittr::extract(1:nrow(germinated))}, # clean up last rep
#       bed=str_sub(block, 1,1)
#       )


germinated$block <- rep(germblocks, length.out=nrow(germinated))
#for low rep mfs, randomize the whole pop
set.seed(777)
germinated_lowrep <- germinated %>%
  group_by(parent.pop, mf) %>%
  mutate(reps=n()) %>%
  filter(reps < 10) %>%
  group_by(parent.pop) %>%
  mutate(block= {map(1:ceiling(n()/10), ~ sample(germblocks)) %>%
       unlist() %>% magrittr::extract(1:n())}, # clean up last rep
         bed=str_sub(block, 1,1)) %>%
  ungroup() 

#For high rep mfs, just use the sequentially assigned blocks
germinated_highrep <- germinated %>%
  group_by(parent.pop, mf) %>%
  mutate(reps=n()) %>%
  filter(reps >= 10) %>%
  ungroup %>%
  mutate(bed=str_sub(block, 1,1))

germinated <- bind_rows(germinated_lowrep, germinated_highrep) %>% arrange(parent.pop, mf)  


# set.seed(777)
# germinated <- germinated %>%
#   arrange(parent.pop, mf) %>%
#   mutate(block= rep(germblocks, length.out = 757)) %>%
#   group_by(parent.pop) %>%
#   mutate(block=sample(block), #sampling within pop
#          bed=str_sub(block, 1,1)) %>%
#   ungroup()


germinated  
```
reality check distribution across mfs
```{r}
germinated %>%
  pivot_wider(id_cols = c(parent.pop, mf), names_from = block, values_from = `germinated?`, values_fill = 0, values_fn = sum)
```

reality check distribution across parent.pops
```{r}
germinated %>%
  pivot_wider(id_cols = parent.pop, names_from = block, values_from = `germinated?`, values_fill = 0, values_fn = sum)
```

plants per block:
```{r}
germinated %>%
  group_by(block) %>%
  summarise(n=n())
```
## assign blocks to nongerminated

We aren't using buffer rows, so can plant 176  plants per bed
```{r}
nrow(notgerminated) 
nrow(notgerminated) / 176
nrow(notgerminated) %% 176

```

So, use 8 beds.  Six seeds won't make it into this design.

get a histogram of rep count

```{r}
notgerminated %>%
  group_by(parent.pop, mf) %>%
  summarize(n=n()) %>%
  ggplot(aes(x=n)) +
  geom_histogram(binwidth = 1)
```

```{r}
notgerminated %>%
  group_by(parent.pop) %>%
  summarize(n=n()) %>%
  arrange(n)
```

remove the six seed from mfs with 13 reps
```{r}
remove <- notgerminated %>%
  group_by(parent.pop, mf) %>%
  mutate(reps=n()) %>%
  ungroup() %>%
  filter(reps==13, rep==13) %>% # use pops that have the highest number mfs in this category
  group_by(parent.pop) %>%
  mutate(mfs=n()) %>%
  arrange(desc(mfs)) %>%
  filter(row_number(mf)==sample(row_number(mf), size = 1)) %>% # pick a random mf for each pop
  ungroup() %>%
  slice_max(mfs, n=6, with_ties = FALSE) %>% # 6 pops with largest number of mfs with 13 non germinated
select(-reps, -mfs)

notgerminated <- notgerminated %>% anti_join(remove)

dim(notgerminated)
```


```{r}
set.seed(777)
notgermblocks <- str_c(rep(notgermbeds,each=2), c("1", "2")) %>% sample

notgermreps <- ceiling(nrow(notgerminated) / length(notgermblocks) ) # number of times we have to assign blocks

# I'm going to be a little more simplistic about the randomization for the non germinated since we can't cope with unequal block size

notgerminated$block <- rep(notgermblocks, length.out=nrow(notgerminated))
#for low rep mfs, randomize the whole pop
set.seed(777)
notgerminated_lowrep <- notgerminated %>%
  group_by(parent.pop, mf) %>%
  mutate(reps=n()) %>%
  filter(reps < 8) %>%
  group_by(parent.pop) %>%
  mutate(block= sample(block),
         bed=str_sub(block, 1,1)) %>%
  ungroup() 

#For high rep mfs, just use the sequentially assigned blocks
notgerminated_highrep <- notgerminated %>%
  group_by(parent.pop, mf) %>%
  mutate(reps=n()) %>%
  filter(reps >= 8) %>%
  ungroup %>%
  mutate(bed=str_sub(block, 1,1))

notgerminated <- bind_rows(notgerminated_lowrep, notgerminated_highrep) %>% arrange(parent.pop, mf)  

notgerminated  
```
reality check distribution across mfs
```{r}
notgerminated %>%
  mutate(notgerminated=1) %>%
  pivot_wider(id_cols = c(parent.pop, mf), names_from = block, values_from = notgerminated, values_fill = 0, values_fn = sum) %>%
  relocate(sort(notgermblocks), .after=mf)
```

reality check distribution across parent.pops
```{r}
notgerminated %>%
    mutate(notgerminated=1) %>%
  pivot_wider(id_cols = parent.pop, names_from = block, values_from = notgerminated, values_fill = 0, values_fn = sum) %>%
    relocate(sort(notgermblocks), .after=parent.pop)
```

plants per block:
```{r}
notgerminated %>%
  group_by(block) %>%
  summarise(n=n())
```
plants per bed
```{r}
notgerminated %>%
  group_by(bed) %>%
  summarise(n=n())
```

Looks good

## planting grid

60 inch veg crops bed is 100 cm of plantable width.  

Create grid
```{r}
stagger <- TRUE
buffer.beds <- 2
buffer.rows <- 2 # on each edge of each bed
exp.beds <- 13
beds <- LETTERS[1:(buffer.beds+exp.beds)]
columns <- 4 # per bed
rows <- 44 # four of which are buffers
plants <- length(beds) * columns * rows
bedwidth <- 100 # cm
furrow <- 52 #cm
size <- 30 # plant diameter
radius <- size/2 
blankevery <- 2 # what is the spacing between blank beds?  (2 means every other; 3 means every third)
expandfactor <- blankevery/(blankevery-1)

plan3 <- expand_grid(bed=beds,
                     column=1:columns,
                     row=1:rows,
                     radius=radius)

plan3 <- plan3 %>%
  mutate(plantedcol=(match(bed, LETTERS)-1)*4 + column,
         bedcol = floor((match(bed, LETTERS)-1)*expandfactor + 1   )) 

plan3
```

add positions
```{r}
if (stagger) {
  ## first part needs to be updated for beds
  column_offset <- sqrt((2*radius)^2 - radius^2) # Pythagorean theorem for offset spacing
  plan3 <- plan3 %>%
    mutate(y_pos=ifelse(is.even(column),
                        row*size, 
                        row*size-radius),
           x_pos=ifelse(column==1,
                        radius+(bedcol-1)*(bedwidth+furrow),
                        radius+(column-1)*column_offset+(bedcol-1)*(bedwidth+furrow)))
} else {
  plan3 <- plan3 %>%
    mutate(y_pos=row*size,
           x_pos=bedcol*(bedwidth+furrow))
}
plan3
```
Plot it
```{r, fig.width=12, eval=FALSE}

#create bed labels
# bed.labels <- plan3 %>%
#   group_by(bed) %>%
#   summarize(x_pos = mean(x_pos+bed_x_offset), y_pos=max(y_pos*1.075)) 


plan3 %>% #filter(bed=="A", row <6) %>%
  ggplot(aes(x0=x_pos, y0=y_pos, r=radius)) +
  geom_circle(fill="lightgreen", alpha=.25) + 
  coord_equal() + 
  #  geom_text(aes(x=x_pos, y=y_pos, label=bed), size= 14, data = bed.labels, inherit.aes = FALSE) + 
  theme(axis.text = element_text(size=18), axis.title = element_text(size=18)) +
  xlab("position (cm)") + ylab("position (cm)")

ggsave("../output/FieldPlan.jpeg", width = 10, height = 6)
```

Plot it in feet
```{r, fig.width=12, eval=FALSE}

#create bed labels
# bed.labels <- plan3 %>%
#   group_by(bed) %>%
#   summarize(x_pos = mean(x_pos+bed_x_offset), y_pos=max(y_pos*1.075)) 


plan3 %>% #filter(bed=="A", row <6) %>%
  mutate(across(.cols=c(x_pos, y_pos, radius), ~ .x*0.0328084)) %>% # convert cm to feet
  ggplot(aes(x0=x_pos, y0=y_pos, r=radius)) +
  geom_circle(fill="lightgreen", alpha=.25) + 
  coord_equal(xlim = c(4,145)) + 
  #  geom_text(aes(x=x_pos, y=y_pos, label=bed), size= 14, data = bed.labels, inherit.aes = FALSE) + 
  theme(axis.text = element_text(size=18), axis.title = element_text(size=18)) +
  xlab("position (feet)") + ylab("position (feet)") 

# ggsave("../output/FieldPlan.jpeg", width = 10, height = 6)
```

Output spacing for Black A to use as guide for cuting holes
```{r, fig.width=8, fig.height=10.5, eval=FALSE}

x_labels <- plan3 %>% filter(bed=="A") %>% pull(x_pos) %>% round(1) %>% unique()

plan3 %>% filter(bed=="A") %>%
  ggplot(aes(x0=x_pos, y0=y_pos, r=radius, label=y_pos)) +
  geom_circle(fill="lightgreen", alpha=.25, color="grey70") + 
  geom_text(aes(x=x_pos, y=y_pos), size=3) +
  scale_x_continuous(breaks = x_labels) +
  coord_equal(ylim = c(20,660)) + 
  #  geom_text(aes(x=x_pos, y=y_pos, label=bed), size= 14, data = bed.labels, inherit.aes = FALSE) + 
  theme(axis.text.x = element_text(angle=90, size=12, vjust = 0.5)) +
  xlab("position (cm)") + ylab("position (cm)") 
ggsave("../output/spacing_guide_1.pdf", width = 8, height = 10.5)

plan3 %>% filter(bed=="A") %>%
  ggplot(aes(x0=x_pos, y0=y_pos, r=radius, label=y_pos)) +
  geom_circle(fill="lightgreen", alpha=.25, color="grey70") + 
  geom_text(aes(x=x_pos, y=y_pos), size=3) +
  scale_x_continuous(breaks = x_labels) +
  coord_equal(ylim = c(700,1320)) + 
  #  geom_text(aes(x=x_pos, y=y_pos, label=bed), size= 14, data = bed.labels, inherit.aes = FALSE) + 
  theme(axis.text.x = element_text(angle=90, size=12, vjust = 0.5)) +
  xlab("position (cm)") + ylab("position (cm)") 
ggsave("../output/spacing_guide_2.pdf", width = 8, height = 10.5)
```


## Assign Plants

### germinated

```{r}
set.seed(999)
germinated <- germinated %>% 
  group_by(bed) %>%
  arrange(block) %>%
  mutate(row=rep(3:(2+ceiling(n()/columns)), each=columns, length.out=n())) %>%
  group_by(block) %>%
  mutate(row=ifelse(str_sub(block,2,2) == "2" & (length(row[row==min(row)]) < columns), row+1, row)) %>%
  mutate(row=sample(row)) %>%
  group_by(bed, row) %>%
  mutate(column=sample(LETTERS[1:columns], size=n())) %>%
  select(-reps)

germinated  %>% arrange(block, row) 

germinated  %>% arrange(parent.pop, mf) # confirm different positions per block
```
Fill in the blanks...
```{r}
germ.fill <- germinated %>% 
  filter(str_detect(block, "1")) %>%
  group_by(block) %>%
  expand(row=1:max(row), column) %>%
  mutate(bed=str_sub(block, 1,1))

germ.fill <- germinated %>%
  filter(str_detect(block, "2")) %>%
  group_by(block) %>%
  expand(row=min(row):(max(row)+2), column) %>%
  bind_rows(germ.fill) %>%
  mutate(bed=str_sub(block, 1,1))

germinated <- germ.fill %>%
  left_join(germinated) %>%
  mutate(parent.pop=ifelse(is.na(parent.pop), "buffer", parent.pop))
         
         # ,
         # block=ifelse(!is.na(block), block, 
         #              ifelse(row < 3, str_c(bed,"1"), str_c(bed, "2"))))

germinated
```


### assign positions for not germinated

```{r}
set.seed(999)
notgerminated <- notgerminated %>% 
  group_by(bed) %>%
  arrange(block) %>%
  mutate(row=rep(1:rows, each=columns, length.out=n())) %>% 
  group_by(block) %>%
  mutate(row=sample(row)) %>%
  group_by(bed, row) %>%
  mutate(column=sample(LETTERS[1:columns], size=n())) %>%
  select(-reps)

notgerminated  %>% arrange(block, row) 
```
combine it
```{r}
plants <- bind_rows(germinated, notgerminated)
```


Write it out
```{r}
plants %>% arrange(parent.pop, mf, rep) %>%
  select(parent.pop, mf, rep, bed, block, row, column, everything() ) %>% write_sheet("https://docs.google.com/spreadsheets/d/1EaSA2QrlWAjFR35pedcvheQp3X7Agsla6pZtE2C1QJ4",
                                                        sheet="New Bed Map")
```

### Correct positions for field mistakes / swaps

```{r}
# Fix replacement made in screenhouse:
cor1 <- plants %>% filter(parent.pop=="YO8" & mf==7 & rep==12) %>%
  mutate(parent.pop="SQ3", mf=2, rep=14)

corrected <- plants %>%
  rows_update(cor1, by=c("block", "row", "column"))

# Fix swaps made while planting

cor2 <- plants %>% filter((parent.pop=="YO8" & mf==2 & rep==8) |( parent.pop=="FR" & mf==5 & rep==1)) %>%
  mutate(row=rev(row))

cor3 <- plants %>% filter((parent.pop=="YO7" & mf==4 & rep==5) | (parent.pop=="TM2" & mf==7 & rep==9)) %>%
  mutate(row=rev(row), column=rev(column))

cor4 <- plants %>% filter((parent.pop=="SQ2" & mf==1 & rep==6) | (parent.pop=="LV1" & mf==1 & rep==6)) %>%
  mutate(row=rev(row), column=rev(column))

cor5 <- plants %>% filter((parent.pop=="YO4" & mf==3 & rep==12) | (parent.pop=="LV1" & mf==7 & rep==10)) %>%
  mutate(row=rev(row), column=rev(column))

corrected <- corrected %>%
  rows_update(bind_rows(cor2, cor3, cor4, cor5), by=c("parent.pop", "mf", "rep"))

# Fix Bed K block swaps

cor6 <- plants %>% filter(block=="K2" & row > 22 & row < 37) %>%
  mutate(block="K1", row=row-22)

cor7 <- plants %>% filter(block=="K1" & row < 15) %>%
  mutate(block="K2", row=row+22)

corrected <- corrected %>%
  rows_update(bind_rows(cor6,cor7), by=c("parent.pop", "mf", "rep"))

plants <- corrected

plants %>% arrange(parent.pop, mf, rep) %>%
  select(parent.pop, mf, rep, bed, block, row, column, everything() ) %>% write_sheet("https://docs.google.com/spreadsheets/d/1EaSA2QrlWAjFR35pedcvheQp3X7Agsla6pZtE2C1QJ4",
                                                        sheet="Corrected Bed Map 2022-12-06")
  
```



## combine bed map and field map

```{r, fig.width=12}

plan3_plants <- plan3 %>%
  mutate(column=LETTERS[column]) %>%
  full_join(plants) %>%
  mutate(label=ifelse(parent.pop=="buffer", "buff", str_c(parent.pop, "\n", mf, "-", rep) )) %>%
  mutate(label=ifelse(is.na(label), "buff", label)) %>%
  filter(!is.na(block))


pdf("../output/UCD2022_planting_maps.pdf",width = 8.5, height = 11)

for(b in unique(plan3_plants$block)) {
  germlabel <- ifelse(b %in% germblocks, " germinated", " NOT germinated")
  pl <- plan3_plants %>% filter(block==b) %>%
    ggplot(aes(x0=x_pos, y0=y_pos, r=radius, label=label)) +
    geom_circle(fill="lightgreen", alpha=.25, color="grey70") + 
    geom_text(aes(x=x_pos, y=y_pos), size=3.5, lineheight=.8) +
    coord_equal() + 
    theme(axis.text.x = element_text(angle=90, size=12, vjust = 0.5)) +
    xlab("position (cm)") + ylab("position (cm)") +
    scale_y_reverse() + 
    ggtitle(str_c("Block ", b, germlabel))
  print(pl)
}

dev.off()

pdf("../output/UCD2022_randomization_maps.pdf",width = 8.5, height = 11)

for(b in unique(plan3_plants$block)) {
  germlabel <- ifelse(b %in% germblocks, " germinated", " NOT germinated")
  pl <- plan3_plants %>% filter(block==b) %>%
    ggplot(aes(x0=column, y0=row, label=label)) +
    geom_label(aes(x=column, y=row), size=3.5, lineheight=.8) +
    coord_equal() + 
    theme(axis.text.x = element_text(angle=90, size=12, vjust = 0.5)) +
    scale_y_reverse(breaks = 1:44) +
    ggtitle(str_c("Block ", b, germlabel))
  print(pl)
}

dev.off()

```

