---
title: "2023 UCD F1s F2s Design"
output: html_notebook
---

## Intro

Determine randomization strategy.

This is written for UCD Fall 2023 planting.

```{r}
library(tidyverse)
library(googlesheets4)
library(ggforce)

is.even <- function(x) x%%2 == 0

url <- "https://docs.google.com/spreadsheets/d/1BM5wVLOffxUdHC1yCBJbv3le5hf7Q-WspYVM_nYBiXQ"
```

## get germination data

parents
```{r}
parents <- read_sheet(url, 
                      sheet = "parent pop labels and germination",
                      na=c("NA", "")
) %>% 
  filter(germinated.y.n == 1) %>%
  select(unique.ID, mf, rack, pop.id, rep = mf.rep) %>%
  mutate(pop.id = ifelse(pop.id %in% c("TM2", "WL2"), #because we have more reps of the mfs for these two it makes sense to randomize per mf instead of per pop.
                         str_c(pop.id, "-", mf),
                         pop.id))
parents
```

F1s
```{r}
F1s <- read_sheet(url, 
                  sheet = "F1s labels and germination",
                  na=c("NA", "")
) %>% 
  filter(germinated.y.n == 1) %>%
  select(unique.ID, rack, pop.id, rep)
F1s
```

F2s
```{r}
F2s <- read_sheet(url, 
                  sheet = "F2s labels and germination",
                  na=c("NA", "")
) %>% 
  filter(germinated.y.n == 1) %>%
  select(unique.ID, rack, pop.id, rep)
F2s
```


```{r}
# extra plants
extra <- read_sheet(url, 
                    sheet = "extra plants",
                    na=c("NA", "")
)  %>%
  mutate(across(.cols= c(`dame-mf`, `sire-mf`), .fns = str_remove, "( |-)[^)]*$" ),
         pop.id = str_c(`dame-mf`, " x ", `sire-mf` ),
         unique.ID=seq(from=2001, length.out=n()),
         rack = NaN) %>%
   select(-`dame-mf`, -`sire-mf`, rep)

extra
```

```{r}
combined <- bind_rows(parents, F1s, F2s, extra)

combined
```

## assign blocks to combined

Can plant 176  plants per bed (44 rows * 4 columns)
```{r}
nrow(combined) 
nrow(combined) / 176
```

get a histogram of rep count

```{r}
combined %>%
  group_by(pop.id) %>%
  summarize(n=n()) %>%
  ggplot(aes(x=n)) +
  geom_histogram(binwidth = 1)
```

```{r}
combined %>%
  group_by(pop.id) %>%
  summarize(n=n()) %>%
  arrange(n)
```

A bit challenging because for pops with few reps total we should be careful to distribute the pops across the blocks.

I think one way this can work is to arrange by pops and then assign a randomized set of block labels to it.

```{r}

beds <- LETTERS[1:6]

set.seed(999)
blocks <- str_c(rep(beds,2), rep(1:2, each=length(beds))) %>% sample() # random arrangement...will be helpful below

reps <- ceiling(nrow(combined) / length(blocks) ) # number of times we have to assign blocks

combined <- combined %>%
  mutate(block = rep(blocks, length.out=n()),
         bed=str_sub(block, 1, 1))

combined  
```
reality check distribution across pop.ids
```{r}
combined %>%
  arrange(block) %>%
  pivot_wider(id_cols = c(pop.id), names_from = block, values_from = bed, values_fill = 0, values_fn = ~ 1)
```

plants per block:
```{r}
combined %>%
  group_by(block) %>%
  summarise(n=n())
```

## planting grid

60 inch veg crops bed is 100 cm of plantable width.  

Create grid
```{r}
stagger <- TRUE
buffer.beds <- 0
buffer.rows <- 1 # on each edge of each bed
exp.beds <- 6
beds <- LETTERS[1:(buffer.beds+exp.beds)]
columns <- 4 # per bed
rows <- 45 # two of which are buffers
plants <- length(beds) * columns * rows
bedwidth <- 100 # cm
furrow <- 52 #cm
size <- 30 # plant diameter in cm
radius <- size/2 

#blankevery wouldn't be needed under current design, but is being kept because the code still uses it.
blankevery <- 2 # what is the spacing between blank beds?  (2 means every other; 3 means every third)

expandfactor <- blankevery/(blankevery-1)

plan3 <- expand_grid(bed=beds,
                     column=1:columns,
                     row=1:rows,
                     radius=radius)

plan3 <- plan3 %>%
  mutate(plantedcol=(match(bed, LETTERS)-1)*4 + column,
         bedcol = floor((match(bed, LETTERS)-1)*expandfactor + 1   )) 

plan3

#note plantedcol is needed for ggplot to get the "columns" in the correct place on the graph.
```

add positions (e.g. the centimeter position of each plant, not just its row and column#)
```{r}
if (stagger) {
  ## first part needs to be updated for beds
  column_offset <- sqrt((2*radius)^2 - radius^2) # Pythagorean theorem for offset spacing
  plan3 <- plan3 %>%
    mutate(y_pos=ifelse(is.even(column),
                        row*size, 
                        row*size-radius),
           x_pos=ifelse(column==1,
                        radius+(bedcol-1)*(bedwidth+furrow),
                        radius+(column-1)*column_offset+(bedcol-1)*(bedwidth+furrow)))
} else { #no stagger
  plan3 <- plan3 %>%
    mutate(y_pos=row*size,
           x_pos=bedcol*(bedwidth+furrow))
}
plan3
```
Plot it
```{r, fig.width=12, eval=FALSE}

#create bed labels
# bed.labels <- plan3 %>%
#   group_by(bed) %>%
#   summarize(x_pos = mean(x_pos+bed_x_offset), y_pos=max(y_pos*1.075)) 


plan3 %>% #filter(bed=="A", row <6) %>%
  ggplot(aes(x0=x_pos, y0=y_pos, r=radius)) +
  geom_circle(fill="lightgreen", alpha=.25) + 
  coord_equal() + 
  #  geom_text(aes(x=x_pos, y=y_pos, label=bed), size= 14, data = bed.labels, inherit.aes = FALSE) + 
  theme(axis.text = element_text(size=18), axis.title = element_text(size=18)) +
  xlab("position (cm)") + ylab("position (cm)")

ggsave("../output/UCD2023-FieldPlan.jpeg", width = 10, height = 6)
```

Plot it in feet
```{r, fig.width=12, eval=FALSE}

#create bed labels
# bed.labels <- plan3 %>%
#   group_by(bed) %>%
#   summarize(x_pos = mean(x_pos+bed_x_offset), y_pos=max(y_pos*1.075)) 


plan3 %>% #filter(bed=="A", row <6) %>%
  mutate(across(.cols=c(x_pos, y_pos, radius), ~ .x*0.0328084)) %>% # convert cm to feet
  ggplot(aes(x0=x_pos, y0=y_pos, r=radius)) +
  geom_circle(fill="lightgreen", alpha=.25) + 
  coord_equal(xlim = c(4,145)) + 
  #  geom_text(aes(x=x_pos, y=y_pos, label=bed), size= 14, data = bed.labels, inherit.aes = FALSE) + 
  theme(axis.text = element_text(size=18), axis.title = element_text(size=18)) +
  xlab("position (feet)") + ylab("position (feet)") 

# ggsave("../output/FieldPlan.jpeg", width = 10, height = 6)
```

Output spacing for Black A to use as guide for cuting holes
```{r, fig.width=8, fig.height=10.5, eval=FALSE}

x_labels <- plan3 %>% filter(bed=="A") %>% pull(x_pos) %>% round(1) %>% unique()

plan3 %>% filter(bed=="A") %>%
  ggplot(aes(x0=x_pos, y0=y_pos, r=radius, label=y_pos)) +
  geom_circle(fill="lightgreen", alpha=.25, color="grey70") + 
  geom_text(aes(x=x_pos, y=y_pos), size=3) +
  scale_x_continuous(breaks = x_labels) +
  coord_equal(ylim = c(20,660)) + 
  #  geom_text(aes(x=x_pos, y=y_pos, label=bed), size= 14, data = bed.labels, inherit.aes = FALSE) + 
  theme(axis.text.x = element_text(angle=90, size=12, vjust = 0.5)) +
  xlab("position (cm)") + ylab("position (cm)") 
ggsave("../output/spacing_guide_1.pdf", width = 8, height = 10.5)

plan3 %>% filter(bed=="A") %>%
  ggplot(aes(x0=x_pos, y0=y_pos, r=radius, label=y_pos)) +
  geom_circle(fill="lightgreen", alpha=.25, color="grey70") + 
  geom_text(aes(x=x_pos, y=y_pos), size=3) +
  scale_x_continuous(breaks = x_labels) +
  coord_equal(ylim = c(700,1320)) + 
  #  geom_text(aes(x=x_pos, y=y_pos, label=bed), size= 14, data = bed.labels, inherit.aes = FALSE) + 
  theme(axis.text.x = element_text(angle=90, size=12, vjust = 0.5)) +
  xlab("position (cm)") + ylab("position (cm)") 
ggsave("../output/spacing_guide_2.pdf", width = 8, height = 10.5)
```


## Assign Plants

### combined

```{r}
set.seed(999)
first_row <- 2 # set to 1 more than number of initial buffer rows
combined <- combined %>% 
  group_by(bed) %>%
  arrange(block) %>%
  mutate(row=rep(first_row:(first_row - 1 + ceiling(n()/columns)), each=columns, length.out=n())) %>%
  group_by(block) %>%
  mutate(row=ifelse(str_sub(block,2,2) == "2" #if block 2 in each bed
                    & (length(row[row==min(row)]) < columns), #and if the first row in block 2 is incomplete
                    row+1, row)) %>% # adjusts for incomplete rows in first block in each bed
  mutate(row=sample(row)) %>% # randomize rows
  group_by(bed, row) %>%
  mutate(column=sample(LETTERS[1:columns], size=n())) # randomize columns

combined  %>% arrange(block, row) 

combined  %>% arrange(pop.id, block) # confirm different positions per block
```

Fill in the blanks...
```{r}
combined.fill <- combined %>% 
  filter(str_detect(block, "1")) %>%
  group_by(block) %>%
  expand(row=1:max(row), column) %>%
  mutate(bed=str_sub(block, 1,1))

combined.fill <- combined %>%
  filter(str_detect(block, "2")) %>%
  group_by(block) %>%
  expand(row=min(row):(max(row)+buffer.rows), column) %>%
  bind_rows(combined.fill) %>%
  mutate(bed=str_sub(block, 1,1))

combined <- combined.fill %>%
  left_join(combined) %>%
  mutate(pop.id=ifelse(is.na(pop.id), "buffer", pop.id))

# ,
# block=ifelse(!is.na(block), block, 
#              ifelse(row < 3, str_c(bed,"1"), str_c(bed, "2"))))

combined
```


Write it out
```{r}
combined %>% arrange(unique.ID) %>%
  select(rack, unique.ID, pop.id, mf, rep, bed, block, row, column, everything() ) %>% write_sheet(url,
                                                                                      sheet="New Bed Map")
```





## combine bed map and field map

```{r, fig.width=12}

plan3_plants <- plan3 %>%
  mutate(column=LETTERS[column]) %>%
  full_join(combined) %>%
  mutate(label=ifelse(pop.id=="buffer", "buff", unique.ID) ) %>%
  mutate(label=ifelse(is.na(label), "buff", label)) %>%
  filter(!is.na(block))


pdf("../output/UCD2023_planting_maps.pdf",width = 8.5, height = 11)

for(b in unique(plan3_plants$block)) {
  pl <- plan3_plants %>% filter(block==b) %>%
    ggplot(aes(x0=x_pos, y0=y_pos, r=radius, label=label)) +
    geom_circle(fill="lightgreen", alpha=.25, color="grey70") + 
    geom_text(aes(x=x_pos, y=y_pos), size=3.5, lineheight=.8) +
    coord_equal() + 
    theme(axis.text.x = element_text(angle=90, size=12, vjust = 0.5)) +
    xlab("position (cm)") + ylab("position (cm)") +
    scale_y_reverse() + 
    ggtitle(str_c("Block ", b))
  print(pl)
}

dev.off()

pdf("../output/UCD2023_randomization_maps.pdf",width = 8.5, height = 11)

for(b in unique(plan3_plants$block)) {
  pl <- plan3_plants %>% filter(block==b) %>%
    ggplot(aes(x0=column, y0=row, label=label)) +
    geom_label(aes(x=column, y=row), size=3.5, lineheight=.8) +
    coord_equal() + 
    theme(axis.text.x = element_text(angle=90, size=12, vjust = 0.5)) +
    scale_y_reverse(breaks = 1:45) +
    ggtitle(str_c("Block ", b))
  print(pl)
}

dev.off()

```

