---
title: "Common Garden Design"
output: html_notebook
---

## Intro

Determine randomization strategy and plant placing for 2024 Wrights Lake 2 Planting

```{r}
library(tidyverse)
library(googlesheets4)
library(ggforce)

is.even <- function(x) x%%2 == 0

popsheet <- "https://docs.google.com/spreadsheets/d/1EhkPTCMpy_hIz2yROdXNtTdh8ljkHjCHnTPdML49MuM"
```
## get germination data

__Note: be sure to check the checkbox to grant access on the webpage that opens when you run the code below__

parents
```{r}
parents <- read_sheet(popsheet, 
                      sheet = "parents_labels_germ",
                      na=c("NA", "")
) %>% 
  filter(germinated.y.n == 1) %>%
  select(unique.ID, rack, pop.id, rep = Rep)
parents
```

F1s
```{r}
F1s <- read_sheet(popsheet, 
                  sheet = "F1s_labels_germ",
                  na=c("NA", "")
) %>% 
  filter(germinated.y.n == 1) %>%
  select(unique.ID, rack, pop.id, rep=Rep)
F1s
```

F2s
```{r}
F2s <- read_sheet(popsheet, 
                  sheet = "F2s_labels_germ",
                  na=c("NA", "")
) %>% 
  filter(germinated.y.n == 1) %>%
  select(unique.ID, rack, pop.id, rep=Rep)
F2s
```


```{r}
# extra plants
# extra <- read_sheet(url, 
#                     sheet = "extra plants",
#                     na=c("NA", "")
# )  %>%
#   mutate(across(.cols= c(`dame-mf`, `sire-mf`), .fns = str_remove, "( |-)[^)]*$" ),
#          pop.id = str_c(`dame-mf`, " x ", `sire-mf` ),
#          unique.ID=seq(from=2001, length.out=n()),
#          rack = NaN) %>%
#    select(-`dame-mf`, -`sire-mf`, rep)
# 
# extra
```

```{r}
#combined <- bind_rows(parents, F1s, F2s, extra)
germinated <- bind_rows(parents, F1s, F2s)
germinated
```

## assign blocks to germinated

get a histogram of rep count

```{r}
germinated %>%
  group_by(pop.id) %>%
  summarize(n=n()) %>%
  ggplot(aes(x=n)) +
  geom_histogram(binwidth = 1)
```

```{r}
germinated %>%
  group_by(pop.id) %>%
  summarize(n=n()) %>%
  arrange(n)
```

A bit challenging because for pops with few reps total we should be careful to distribute the pops across the blocks.  Also different max rep sizes: 16 for F2s, 26 for at least a couple of F1s.

Will go for 16 blocks total.  That will be one rep per block for fully germinated F2s.

I think one way this can work is to arrange by pops and mfs and then assign a randomized set of block labels to it.


```{r}
#terminology with "germ" prefix is a leftover from UCD2022 when we planted both germinated and ungerminated seed.

nblocks <- 16

germblocks <- LETTERS[1:nblocks]

germreps <- ceiling(nrow(germinated) / length(germblocks) ) # number of times we have to assign blocks

germinated$block <- rep(germblocks, length.out=nrow(germinated))

#for low rep mfs, randomize all mfs in the pop that have have < nblocks
set.seed(888)
germinated_low_extra_rep <- germinated %>%
 # group_by(pop.id, mf) %>%
  group_by(pop.id) %>%
  mutate(reps=n()) %>%
  filter(reps < 16 | rep > 16) %>% # because nblocks represents full replication.  Pull pops with less than nblocks reps and pull extra reps from those with > nblocks
  group_by(pop.id) %>%
  mutate(block= {map(1:ceiling(n()/nblocks), ~ sample(germblocks)) %>% #sampling without replacement 1 block at a time helps with even distribution across blocks.
      unlist() %>% magrittr::extract(1:n())}, # clean up last rep
  ) %>%
  ungroup() 

#For high rep mfs, just use the sequentially assigned blocks
germinated_fullrep <- germinated %>%
  #group_by(pop.id, mf) %>%
  group_by(pop.id) %>%
  mutate(reps=n()) %>%
  filter(reps >= nblocks, rep < nblocks+1) %>% #get those plants corresponding to full germination but not extra
  ungroup()

germinated <- bind_rows(germinated_low_extra_rep, germinated_fullrep) %>% 
  #arrange(pop.id, mf)
  arrange(pop.id)

germinated  
```
reality check distribution across mfs
```{r}
# germinated %>%
#   pivot_wider(id_cols = c(pop.id, mf), names_from = block, values_from = `germinated?`, values_fill = 0, values_fn = sum)
```

reality check distribution across pop.ids
```{r}
germinated %>%
  pivot_wider(id_cols = pop.id, names_from = block, values_from = reps, values_fill = 0, values_fn = sum) %>%
  relocate(germblocks)
```

plants per block:
```{r}
germinated %>%
  group_by(block) %>%
  summarise(n=n()) %>% arrange(n)
```

## planting grid

Plot is trapezoidal.  wide edges are 20m and 16.5m.  narrow are 15m and 7.5m .

__numbers below updated to match more precise field measurements__

Try creating rectangular grid and then use blanks to narrow it.

Create grid
```{r}
stagger <- TRUE
buffer.beds <- 0
buffer.rows <- 2 # on each edge of each bed
exp.beds <- 11
beds <- letters[1:(buffer.beds+exp.beds)]
columns <- 4 # per bed
rows <- 63 # four of which are buffers
plants <- length(beds) * columns * rows
bedwidth <- 90 # cm
furrow <- 100 #cm
size <- 25 # plant diameter
radius <- size/2 

plan3 <- expand_grid(bed=beds,
                     column=1:columns,
                     row=1:rows,
                     radius=radius)

plan3 <- plan3 %>%
  mutate(plantedcol=(match(bed, letters)-1)*4 + column,
         bedcol = floor((match(bed, letters)-1) + 1   )) 

plan3
```

add positions
```{r}
if (stagger) {
  column_offset <- sqrt((2*radius)^2 - radius^2) # Pythagorean theorem for offset spacing
  plan3 <- plan3 %>%
    mutate(y_pos=ifelse(is.even(column),
                        row*size, 
                        row*size-radius),
           x_pos=ifelse(column==1,
                        radius+(bedcol-1)*(bedwidth+furrow),
                        radius+(column-1)*column_offset+(bedcol-1)*(bedwidth+furrow)))
} else {
  plan3 <- plan3 %>%
    mutate(y_pos=row*size,
           x_pos=bedcol*(bedwidth+furrow))
}
plan3
```

add indicator for whether or not a position is available and inbounds 

```{r}
bedrows   <- c(63, 61, 60, 59, 53, 47, 45, 34, 24, 20, 16)

plan3 <- plan3 %>%
  mutate(status=ifelse(row <= bedrows[bedcol], "available", "out-of-bounds"))

# eliminate some beds for 2024 because we have fewer plants

plan3 <- plan3 %>%
  mutate(status=ifelse(str_detect(bed, "[abjk]"), "out-of-bounds", status))

# block out tree position (approx)
plan3 <- plan3 %>%
  mutate(status=ifelse(bedcol == 5 & row > bedrows[5] -23 & row <= bedrows[5]-13, "out-of-bounds", status),
         status=ifelse(bedcol == 6 & row > bedrows[6] -19 & row <= bedrows[6]-9, "out-of-bounds", status)
         )

# block out positions with rocks

rocks <- tribble(~bed, ~column, ~row,
                 "a", 1, 44,
                 "a", 2, 44,
                 "a", 3, 45,
                 "a", 3, 45,
                 "a", 2, 48,
                 "a", 4, 48,
                 "b", 1, 35,
                 "c", 4, 3,
                 "c", 4, 4,
                 "c", 1, 9,
                 "c", 3, 26,
                 "c", 4, 26,
                 "c", 3, 27,
                 "c", 4, 27,
                 "c", 1, 33,
                 "c", 3, 45,
                 "c", 3, 46,
                 "d", 2, 5,
                 "d", 1, 6,
                 "d", 3, 6,
                 "d", 4, 6,
                 "d", 1, 7,
                 "d", 2, 7,
                 "d", 4, 7,
                 "d", 2, 8,
                 "d", 1, 12,
                 "d", 2, 34,
                 "d", 3, 37,
                 "e", 1, 6,
                 "e", 2, 9,
                 "e", 2, 13,
                 "e", 4, 15,
                 "e", 4, 16,
                 "e", 1, 30,
                 "e", 2, 30,
                 "e", 3, 30,
                 "e", 4, 30,
                 "f", 2, 5,
                 "f", 3, 6,
                 "f", 4, 17,
                 "f", 4, 18,
                 "f", 2, 40,
                 "g", 3, 23,
                 "g", 4, 22,
                 "g", 4, 23,
                 "i", 1, 3,
                 "i", 3, 4,
                 "i", 4, 5,
                 "k", 3, 5,
                 )

rocks <- bind_rows(rocks,
  tibble(
    bed="h", 
    column = c(rep(1,7),
                rep(2,3),
                rep(3, 6),
                rep(4, 3)),
    row = c(9:12, 14:16,
            7, 14, 16,
            4, 5, 7, 14, 16, 17,
            11, 12, 13)
  )
) 

rocks <- rocks %>% mutate(new_status="out-of-bounds") %>% arrange(bed, row, column)

plan3 <- left_join(plan3, rocks) %>% 
  mutate(status = ifelse(is.na(new_status), status, new_status)) %>%
  select(-new_status)

plan3
```

Get 2023 plants that are still alive
```{r, warning=FALSE}
survey <- read_sheet("https://docs.google.com/spreadsheets/d/1cJE-tR3v4OA5rurcnFU5f1MgtEbxECSre5nzRiVN0-w", na = c("", "NA"),
                     col_types = "ccicciicc") %>%
  filter(str_detect(death.date, "[ABC]"))
```


```{r}
survey <- survey %>%
  select(bed, column=`bed- col`, row=`bed- row`)  %>%
  mutate(bed = tolower(bed),
         new_status="2023-survivor",
         column=case_match(column,
                        "A" ~ 1,
                        "B" ~ 2,
                        "C" ~ 3,
                        "D" ~ 4
                        ))

plan3 <- left_join(plan3, survey) %>%
  mutate(status = ifelse(is.na(new_status), status, new_status)) %>%
  select(-new_status)
```

Get 2023 TM2 plants that fruited

```{r}
annual_census_2023 <- read_csv("~/Library/CloudStorage/GoogleDrive-jnmaloof@ucdavis.edu/Shared drives/IntBioTeam/Common Gardens/WL2_2023/CorrectedCSVs/WL2_annual_census_20231027_corrected.csv", guess_max = 2000) %>%
  filter(pop=="TM2", num.fruit>0) %>%
  select(bed, column=`bed-col`, row=`bed-row`)  %>%
  mutate(bed = tolower(bed),
         new_status="2023-TM2-fruit",
         column=case_match(column,
                        "A" ~ 1,
                        "B" ~ 2,
                        "C" ~ 3,
                        "D" ~ 4
                        ))

plan3 <- left_join(plan3, annual_census_2023) %>%
  mutate(status = ifelse(is.na(new_status), status, new_status)) %>% 
  select(-new_status)

``` 
Add buffers at the beginning and end of each bed
```{r}
plan3 <- plan3 %>%
  
  #top buffers:
  mutate(buffer=(row<=3 & str_detect(bed, "[c-h]")) | (row <=2 & str_detect(bed, "i")), # if you want different number of buffers in last row

  #bottom buffers
         buffer=ifelse(row > (bedrows[bedcol]-2) & row <= bedrows[bedcol] & str_detect(bed, "[c-i]"), TRUE, buffer),
         status=ifelse(buffer, "buffer", status)
  )
```



Plot it
```{r, fig.width=12, eval=TRUE}

plan3 %>%
  ggplot(aes(x0=x_pos, y0=y_pos, r=radius, fill=status)) +
  geom_circle(alpha=.75) + 
  scale_fill_manual(values=c("out-of-bounds"="grey40", "available"="lightgreen", "2023-survivor"="blue", "2023-TM2-fruit"="magenta", "buffer"="yellow")) +
  coord_equal() + 
  #  geom_text(aes(x=x_pos, y=y_pos, label=bed), size= 14, data = bed.labels, inherit.aes = FALSE) + 
  theme(axis.text = element_text(size=18), axis.title = element_text(size=18)) +
  xlab("position (cm)") + ylab("position (cm)") +
  scale_y_reverse() +
  ggtitle(str_c("Total available positions = ", sum(plan3$status=="available")))

ggsave("../output/2024-WL2-FieldPlan.jpeg", width = 10, height = 6)
```

Position blocks in the available spots
```{r}
blocksizes <- germinated %>%
  group_by(block) %>%
  summarise(n=n()) %>%
  mutate(n=n+4) # a buffer row between blocks

blocks <-rep(blocksizes$block, time=blocksizes$n)

avail <- plan3 %>%
  filter(status=="available") %>%
  arrange(bed, row, column) %>%
  magrittr::extract(1:length(blocks),) %>%
  mutate(block=blocks) %>%
  group_by(block) %>%
  mutate(block.id=str_c(block, str_pad(1:n(), width = 2, pad = "0"))) %>% 
  ungroup()


```


## Assign Plants

### germinated

```{r}
set.seed(999)

germinated <- germinated %>% 
  group_by(block) %>%
  slice_sample(n=nrow(.)) %>% # randomize position within blocks
  mutate(block.id=str_c(block, str_pad(1:n(), width = 2, pad = "0"))) %>%
  ungroup() %>%
  arrange(block.id) %>%
  select(-reps)

germinated <- left_join(avail, germinated)

germinated  %>% arrange(block, row, column) %>%
  select(block, row, column, block.id, everything())

germinated  %>% arrange(pop.id) %>%# confirm different positions per block
  select(block, bed, row, column, pop.id)
```



Fill in the blanks..
```{r}
# germ.fill <- germinated %>%
#   group_by(block) %>%
# #  expand(row=1:max(row), column) #no end buffer rows.  But should fill ragged edges.
#    expand(row=1:max(row+1), column) # use this if you want buffer rows at the end of each bed
# 
# germinated <- germ.fill %>%
#   left_join(germinated) %>%
#   mutate(pop.id=ifelse(is.na(pop.id), "buffer", pop.id))
# 
# germinated
```





### Correct positions for field mistakes / swaps

```{r}
## not updated for 2023 WL2
# # Fix replacement made in screenhouse:
# cor1 <- plants %>% filter(parent.pop=="YO8" & mf==7 & rep==12) %>%
#   mutate(parent.pop="SQ3", mf=2, rep=14)
# 
# corrected <- plants %>%
#   rows_update(cor1, by=c("block", "row", "column"))
# 
# # Fix swaps made while planting
# 
# cor2 <- plants %>% filter((parent.pop=="YO8" & mf==2 & rep==8) |( parent.pop=="FR" & mf==5 & rep==1)) %>%
#   mutate(row=rev(row))
# 
# cor3 <- plants %>% filter((parent.pop=="YO7" & mf==4 & rep==5) | (parent.pop=="TM2" & mf==7 & rep==9)) %>%
#   mutate(row=rev(row), column=rev(column))
# 
# cor4 <- plants %>% filter((parent.pop=="SQ2" & mf==1 & rep==6) | (parent.pop=="LV1" & mf==1 & rep==6)) %>%
#   mutate(row=rev(row), column=rev(column))
# 
# cor5 <- plants %>% filter((parent.pop=="YO4" & mf==3 & rep==12) | (parent.pop=="LV1" & mf==7 & rep==10)) %>%
#   mutate(row=rev(row), column=rev(column))
# 
# corrected <- corrected %>%
#   rows_update(bind_rows(cor2, cor3, cor4, cor5), by=c("parent.pop", "mf", "rep"))
# 
# # Fix Bed K block swaps
# 
# cor6 <- plants %>% filter(block=="K2" & row > 22 & row < 37) %>%
#   mutate(block="K1", row=row-22)
# 
# cor7 <- plants %>% filter(block=="K1" & row < 15) %>%
#   mutate(block="K2", row=row+22)
# 
# corrected <- corrected %>%
#   rows_update(bind_rows(cor6,cor7), by=c("parent.pop", "mf", "rep"))
# 
# plants <- corrected
# 
# plants %>% arrange(parent.pop, mf, rep) %>%
#   select(parent.pop, mf, rep, bed, block, row, column, everything() ) %>% write_sheet("https://docs.google.com/spreadsheets/d/1EaSA2QrlWAjFR35pedcvheQp3X7Agsla6pZtE2C1QJ4",
#                                                         sheet="Corrected Bed Map 2022-12-06")

```



## combine bed map and field map

first get block rows correct and write to gsheets
```{r}
germinated <- germinated %>%
  ungroup() %>%
  arrange(block, row, column) %>%
  rename(blockrow=row, blockcolumn=column) %>%
  group_by(block) %>%
  mutate(blockrow=as.numeric(as.factor(blockrow))) %>%
  ungroup()


# germinated %>% arrange(pop.id, mf, rep) %>%
#   select(pop.id, mf, rep, block, row, column, everything() ) %>% 
#   write_sheet(popsheet,
#               sheet="Buffer New Bed Map")

germinated %>% arrange(rack, unique.ID) %>%
  select(rack, unique.ID, pop.id, rep, block, blockrow, blockcolumn ) %>%
  write_sheet(popsheet,
              sheet="Randomization Map")
```


```{r}
plan3 <- plan3 %>%
  ungroup() %>%
  arrange(bedcol, row, column) %>%
  rename(bedrow=row)

#combine
plan3_plants <- full_join(plan3, germinated)

# reverse positions in even beds (to "snake" the blocks through the beds)
# plan3_plants <- plan3_plants %>%
#   group_by(bedcol) %>%
#   mutate(across(.cols=block:rep, .fns= ~ ifelse(bedcol%%2==0, rev(.x), .x))) %>%
#   ungroup()


# set up labels
plan3_plants <- plan3_plants %>%
  mutate(column=LETTERS[column]) %>%
  mutate(#label=ifelse(buffer, "buff", str_c(parent.pop, "\n", mf, "-", rep) ),
    label=ifelse(buffer, "buff", unique.ID),
    label=ifelse(status=="available" & pop.id=="buffer", "buff", label),
    label=ifelse(is.na(label), "", label),
    colorblock=ifelse(is.na(block), status, block),
    colorblock=ifelse(status=="available" & pop.id=="buffer", "buffer", colorblock))

plan3_plants

#Write it out

plan3_plants %>% arrange(pop.id, rep) %>%
  select(pop.id, rep, everything() ) %>%
  write_sheet(popsheet,
              sheet="Buffer New Bed Map")
```


```{r, fig.height=12}
pdf("../output/2024-WL2-planting_maps.pdf",width = 8.5, height = 11)

colors=c("grey40", "yellow", "blue", "magenta", rep(c("orange", "skyblue", "green"), length.out=length(germblocks))) %>%
  set_names("out-of-bounds", "buffer", "2023-survivor", "2023-TM2-fruit", germblocks)


for(b in unique(plan3_plants$bed)) {
  pl1 <- plan3_plants %>% filter(bed==b, bedrow<33) %>%
    ggplot(aes(x0=x_pos, y0=y_pos, r=radius, label=label, fill=colorblock)) +
    scale_fill_manual(values=colors) +
    geom_circle(alpha=.5, color="grey70") +
    geom_text(aes(x=x_pos, y=y_pos), size=3, lineheight=.8) +
    theme(axis.text.x = element_text(angle=90, size=12, vjust = 0.5)) +
    xlab("position (cm)") + ylab("position (cm)") +
    scale_y_reverse() +
    coord_equal(ratio=.4) + 
    ggtitle(str_c("Bed ", b, ", sheet 1"))
  print(pl1)
  pl2 <- plan3_plants %>% filter(bed==b, bedrow>=33) %>%
    ggplot(aes(x0=x_pos, y0=y_pos, r=radius, label=label, fill=colorblock)) +
    scale_fill_manual(values=colors) +
    geom_circle(alpha=.5, color="grey70") +
    geom_text(aes(x=x_pos, y=y_pos), size=3, lineheight=.8) +
    theme(axis.text.x = element_text(angle=90, size=12, vjust = 0.5)) +
    xlab("position (cm)") + ylab("position (cm)") +
    scale_y_reverse() +
    coord_equal(ratio=.4) + 
    ggtitle(str_c("Bed ", b, ", sheet 2"))
  print(pl2)
}

dev.off()
```

Randomization maps

```{r, fig.height=11, fig.width=8.5}

pdf("../output/2024-WL2-randomization_maps.pdf",width = 8.5, height = 11)

germinated <- germinated %>% arrange(block, blockrow, blockcolumn)

for(b in unique(germinated$block)) {
  pl <- germinated %>% filter(block==b) %>%
#    mutate(label=ifelse(parent.pop=="buffer", "buff", str_c(parent.pop, "\n", mf, "-", rep) )) %>%
    mutate(label=ifelse(pop.id=="buffer", "buff", unique.ID)) %>%
    mutate(label=ifelse(is.na(label), "buff", label)) %>%
    ggplot() +
    geom_label(aes(x=blockcolumn, y=blockrow, label=label), size=3, lineheight=.8) +
    theme(axis.text = element_text(size=12)) +
    scale_y_reverse(breaks = 1:31) +
    scale_x_discrete(position = "top") +
    ggtitle(str_c("Block ", b)) +
    coord_equal(ratio=.5)
  print(pl)
}

dev.off()

```


