---
title: "F1s and F2s to Plant"
author: "Brandie Quarles"
date: "`r Sys.Date()`"
output: html_document
---
Final Design:
-   F1s: Only types with > 20 seeds, 26 pots per type, 2 seeds per pot when possible
-   F2s: All types available, 16 pots per type, 3 seeds per pot when possible 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries
```{r}
library(tidyverse)
library(googlesheets4)
url_f1s <- "https://docs.google.com/spreadsheets/d/1t5X39iisjTCknwkyNLs8Q7bmiKYkmkWtPZbpj3Ce4N8/edit#gid=709784644"
url_f2s <- "https://docs.google.com/spreadsheets/d/1bw6rzqCcrGjE5DuIe4oxC_GMPKG66nHHNyT_iY5umBQ/edit#gid=729112464"
```

## Load Data

List of all pops
```{r}
allpops <- c(F1s$pop.id, F2s$pop.id) %>% unique() %>%
  str_extract_all("[A-Z]{2,3}[1-9]{0,2}") %>%
  unlist() %>%
  unique()
allpops
```

### Elevation
```{r}
el <- read_sheet("https://docs.google.com/spreadsheets/d/1FORIRL-1J15fD5iPqVyv_EWbc6uhV83-NL8SsjJL0Hw/edit#gid=0") %>% 
  filter(`Species Code`=="STTO") %>%
  mutate(elevation=unlist(`Elevation (m)`)) %>%
  select(pop.id=`Site code`, elevation) %>%
  mutate(pop.id=str_replace(pop.id, "YOSE", "YO")) %>% 
  filter(!is.na(pop.id))
el

el_high_low <- el %>% 
  mutate(elevation_class=ifelse(elevation>2000, "high", "low"))
el_high_low
```

### F1s
```{r}
F1s <- read_sheet(url_f1s,
                  sheet = "Collected fruits list F1 seed",
                  na=c("NA", "", "unknown")) 
head(F1s)

F1s_short <- F1s %>% select(dame_pop=`dame pop`, dame_mf_rep=`dame mf-rep`, sire_pop=`sire pop`, sire_mf_rep=`sire mf-rep`, date_crossed=`date crossed`, date_collected=`date collected`, seeds_avail=`number of seed updated after fall 2023 planting`, seed_quality=`seed quality`, notes)
head(F1s_short)

F1s_short_summarize <- F1s_short %>% group_by(dame_pop, sire_pop) %>% summarise(Sum_Seeds=sum(seeds_avail))
F1s_short_summarize
dim(F1s_short_summarize) #40 F1 types

F1s_short_summarize_twenty <- F1s_short_summarize %>% 
  filter(Sum_Seeds >= 20) %>%
  filter(!is.na(sire_pop)) %>% 
  arrange(Sum_Seeds)
F1s_short_summarize_twenty
F1s_short_summarize_twenty %>% filter(Sum_Seeds==20) #13 of the 26 F1s with at least 20 seeds only have 20 seeds 

F1s_short_summarize_twenty %>% filter(Sum_Seeds>20) %>% arrange(desc(Sum_Seeds))

F1s_plant <- F1s_short_summarize_twenty %>% filter(Sum_Seeds>20) %>% 
  mutate(plantable_seeds = Sum_Seeds - 20,
         pots_to_plant = ifelse(plantable_seeds<26, plantable_seeds, 26),
         seeds_to_plant = ifelse(plantable_seeds<52, plantable_seeds, 52)
         ) %>% 
  arrange(desc(Sum_Seeds))
F1s_plant
sum(F1s_plant$pots_to_plant)
```

### F2s
```{r}
F2s <- read_sheet(url_f2s,
                  sheet = "Collected fruits F2 seed",
                  #col_types=cols(.default=col_character()), #force all columns to be character
                  na=c("NA", "")) 
head(F2s) #need to make everything a character 

F2s_short <- F2s %>% 
  select(dame_pop=`dame pop/hybrid`, dame_mf_rep=`dame mf-rep/ID`, sire_pop=`sire pop/hybrid`,
         sire_mf_rep=`sire mf-rep/ID`, date_crossed=`date crossed`, 
         date_collected=`date collected`, emasc_poll=`emasulator-pollinator`, 
         seeds_avail=`number of seed updated after fall 2023 planting`, 
         seed_quality=`seed quality`, notes) %>% 
  filter(!is.na(dame_pop)) %>% 
  mutate(date_collected=ymd(date_collected), Year_Collected=year(date_collected),
         Year_Collected=ifelse(is.na(Year_Collected), 2023, Year_Collected))
head(F2s_short)

F2s_short_after_ripened <- F2s_short %>% filter(Year_Collected < 2024)

F2s_plant <- F2s_short_after_ripened %>% 
  group_by(dame_pop, sire_pop) %>% 
  summarise(Unique_Crosses=sum(!is.na(seeds_avail)), Sum_Seeds=sum(seeds_avail)) %>% 
  mutate(pots_to_plant=ifelse(Sum_Seeds < 16, Sum_Seeds, 16),
         seeds_to_plant=ifelse(Sum_Seeds < 48, Sum_Seeds, 48)) %>% 
  mutate(dame_pop=str_replace(dame_pop, "X", "x")) %>% 
  arrange(Sum_Seeds) 
F2s_plant
dim(F2s_plant) #67 types of F2s
sum(F2s_plant$Sum_Seeds) #1523 total seeds available
sum(F2s_plant$pots_to_plant) #920 pots 

F2s_short_after_ripened_summarize_WL2 <- F2s_plant %>% 
  mutate(WL2_mom =str_detect(dame_pop, "WL2"), 
         WL2_dad=str_detect(sire_pop, "WL2"), 
         WL2_cross=ifelse(WL2_mom==TRUE | WL2_dad==TRUE, TRUE, FALSE)) %>% 
  filter(WL2_cross == TRUE)
head(F2s_short_after_ripened_summarize_WL2)
dim(F2s_short_after_ripened_summarize_WL2) #62 types of F2s with WL2 involved 

F2s_short_after_ripened_summarize_TM2 <- F2s_plant %>% 
  mutate(TM2_mom =str_detect(dame_pop, "TM2"), 
         TM2_dad=str_detect(sire_pop, "TM2"), 
         TM2_cross=ifelse(TM2_mom==TRUE | TM2_dad==TRUE, TRUE, FALSE)) %>% 
  filter(TM2_cross == TRUE)
head(F2s_short_after_ripened_summarize_TM2)
dim(F2s_short_after_ripened_summarize_TM2) #33 types of F2s with TM2 involved
```

## Pots per planting 
Anything that’s 100% maternal high elevation should be planted with high elevation populations (8 weeks strat) 
Anything that 100% maternal low should be planted with low elevation populations (4 weeks strat) 
Everything else is planted in the middle (6 weeks strat)

### Parents
```{r}
parents <- tibble(pop.id=c("LV1", "SQ3", "WL1", "YO11", "BH", "CC", "DPR", "WV", "TM2", "WL2"), 
                  pots_to_plant=c(21, 21, 21, 21, 21, 21, 21, 21, 98, 98))
parents

parents_elev <- left_join(parents, el_high_low) %>% arrange(elevation_class)
parents_elev #note for UCD garden, WL1 was planted with high elevation pops 

parents_elev_summary <- parents_elev %>% 
  group_by(elevation_class) %>% 
  summarise(Total_Pots=sum(pots_to_plant))
parents_elev_summary #high 161, low 203
```

### F1s 
```{r}
el_high_low_f1s <- el_high_low %>% select(dame_pop=pop.id, elevation_class)
F1s_dames <- F1s_plant %>% select(dame_pop)
F1s_dames_elev <- left_join(F1s_dames, el_high_low_f1s) %>% distinct()
F1s_dames_elev

F1s_plant_elev <- left_join(F1s_dames_elev, F1s_plant) %>% arrange(elevation_class)
F1s_plant_elev

F1s_plant_elev_summary <- F1s_plant_elev %>% 
  group_by(elevation_class) %>% 
  summarise(Total_Pots=sum(pots_to_plant))
F1s_plant_elev_summary #87 high, 181 low 
```

### F2s
```{r}
el_high_low_dame1 <- el_high_low %>% select(dame1=pop.id, dame1_elev=elevation_class)
el_high_low_dame2 <- el_high_low %>% select(dame2=pop.id, dame2_elev=elevation_class)

F2s_plant_dames <- F2s_plant %>% 
  select(dame_pop) %>% 
  separate(col = dame_pop,
           into = c("dame1", "dame2"),
           sep = " x ",
           remove = FALSE) %>% 
  mutate(dame2=ifelse(is.na(dame2), dame1, dame2)) %>% 
  distinct()
F2s_plant_dames

F2s_dame1 <- left_join(F2s_plant_dames, el_high_low_dame1)
F2s_dame2 <- left_join(F2s_plant_dames, el_high_low_dame2)
F2s_dames_elev <- left_join(F2s_dame1, F2s_dame2) %>% 
  mutate(dame1_elev=ifelse(dame1_elev=="low", 0, 1),
         dame2_elev=ifelse(dame2_elev=="low", 0, 1),
         maternal_high=(dame1_elev+dame2_elev)/2) %>% 
  select(dame_pop, maternal_high)
F2s_dames_elev

F2s_plant_elev <- full_join(F2s_dames_elev, F2s_plant) %>% arrange(maternal_high)
F2s_plant_elev

F2s_plant_elev_summary <- F2s_plant_elev %>% 
  group_by(maternal_high) %>% 
  summarise(Total_Pots=sum(pots_to_plant))
F2s_plant_elev_summary #303 high, 414 middle, 203 low 
```

