---
title: "Common Garden Design"
output: html_notebook
---

## Intro

Determine randomization strategy and plant placing for 2026 Wrights Lake 2 Planting

```{r}
library(tidyverse)
library(googlesheets4)
library(ggforce)

is.even <- function(x) x%%2 == 0

popsheet <- "https://docs.google.com/spreadsheets/d/12E6iegjdVGfiqv1H9B_WJC0qA0_OlAdwDQrbbUI-OP8/edit?gid=852415812#gid=852415812"
```

## IMPORTANT

Snaking blocks through the beds really should be done but it is tricky to coordinate with the blocked positions from the field.

## get germination and transplant data

```{r}
germ_data <- read_csv("../input/WL2_2026_InitialSize_DNA_corrected_inprogress.csv") %>% 
  select(Rack:Unique.ID, germ.y.n)

popinfo <- read_sheet(popsheet, 
                  sheet = "Labeling-ToPrint",
                  na=c("NA", ""),
                  col_types = "c") %>% 
  select(Type, pop.id:sire_mf, Rep, Unique.ID) %>% 
  mutate(Unique.ID=as.numeric(Unique.ID))
popinfo_norep <- popinfo %>% select(-Rep)

transplants <- read_csv("../input/WL2_2026_Transplanting_corrected.csv") #transplant info (extra germs)
transplants_mf <- transplants %>% 
  select(-rep) %>% 
  mutate(germ.y.n="Y") %>% 
  left_join(popinfo_norep, by=join_by(old_uniquelD == Unique.ID, pop.id)) %>% 
  select(Type, pop.id, dame_mf:sire_mf, Unique.ID=new_uniquelD, Rack=new_rack, germ.y.n)

newsdls <- tibble(old_uniquelD=c(3252, 3252, 3269, 3269, 3338, 3338, 3380, 3396),
                  new_uniquelD=c(3953, 3954, 3955, 3956, 3957, 3958, 3959, 3960),
                  germ.y.n="Y") #add sdls found during size/DNA census 
newsdlspops <- newsdls %>% 
  left_join(popinfo_norep, by=join_by(old_uniquelD == Unique.ID)) %>% 
  select(Type, pop.id, dame_mf:sire_mf, Unique.ID=new_uniquelD, germ.y.n)


germ_only <- germ_data %>% 
  left_join(popinfo) %>% 
  filter(!is.na(pop.id)) %>% 
  bind_rows(transplants_mf) %>% 
  bind_rows(newsdlspops) %>% 
  rename(germinated.y.n=germ.y.n) %>% 
  #remove plants that were not there during size/DNA
  filter(Unique.ID!=3038) %>% 
  filter(Unique.ID!=3490) %>% 
  filter(Unique.ID!=3514) %>% 
  filter(Unique.ID!=3920) %>% 
  filter(Unique.ID!=3815) %>% 
  filter(germinated.y.n=="Y")
```

```{r}
germinated <- germ_only %>% 
  unite("mf", dame_mf:sire_mf, remove = FALSE, na.rm = TRUE) %>% 
  group_by(pop.id) %>% 
  mutate(rep_for_blocking=row_number()) %>%  #add smooth rep number for block assignments
  mutate(reps=n()) #sum number of reps per pop
#605 plants germinated
names(germinated)
germinated %>% arrange(pop.id) %>% head()
```

## assign blocks to germinated

get a histogram of rep count

```{r}
germinated %>%
  group_by(pop.id) %>%
  summarize(n=n()) %>%
  ggplot(aes(x=n)) +
  geom_histogram(binwidth = 1)

summary(germinated)
```

```{r}
germinated %>%
  group_by(pop.id) %>%
  summarize(n=n()) %>%
  arrange(n)

germinated %>%
  group_by(pop.id) %>%
  summarize(n=n()) %>% 
  summary()
```

A bit challenging because for pops with few reps total we should be careful to distribute the pops across the blocks.\
Also different max rep sizes: \~30 for a couple BC1 (With one >90), 5 for the BC2.

Will go for 16 blocks total (that's the median sample size)

I think one way this can work is to arrange by pops and then assign a randomized set of block labels to it.

```{r}
#terminology with "germ" prefix is a leftover from UCD2022 when we planted both germinated and ungerminated seed.

nblocks <- 16

germblocks <- LETTERS[1:nblocks]

germreps <- ceiling(nrow(germinated) / length(germblocks) ) # number of times we have to assign blocks

germinated$block <- rep(germblocks, length.out=nrow(germinated))

#for low rep mfs, randomize all mfs in the pop that have have < nblocks
set.seed(888)
germinated_low_extra_rep <- germinated %>%
  group_by(pop.id) %>%
  filter(reps < 16 | rep_for_blocking > 16) %>% # because nblocks represents full replication.  
  #Pull pops with less than nblocks reps and pull extra reps from those with > nblocks
  group_by(pop.id) %>%
  mutate(block= {map(1:ceiling(n()/nblocks), ~ sample(germblocks)) %>% #sampling without replacement 1 block at a time helps with even distribution across blocks.
      unlist() %>% magrittr::extract(1:n())}, # clean up last rep
  ) %>%
  ungroup() 

#For high rep pops, just use the sequentially assigned blocks
germinated_fullrep <- germinated %>%
  group_by(pop.id) %>%
  filter(reps >= nblocks, rep_for_blocking < nblocks+1) %>% #get those plants corresponding to full germination but not extra
  ungroup()

germinated_blocks <- bind_rows(germinated_low_extra_rep, germinated_fullrep) %>% 
  arrange(pop.id)

summary(germinated_blocks)
```

### reality check distribution across pop.ids

```{r}
germinated_blocks %>%
  group_by(pop.id, block) %>% 
  mutate(reps_per_block=n()) %>% 
  select(pop.id, block, reps_per_block) %>%
  distinct() %>% 
  select(pop.id, block, reps_per_block) %>% 
  pivot_wider(names_from = block, values_from = reps_per_block, values_fill=0) %>%
  relocate(germblocks)
```

### plants per block:

```{r}
germinated_blocks %>%
  group_by(block) %>%
  summarise(n=n()) %>% arrange(n)
```

## planting grid

Plot is trapezoidal. wide edges are 20m and 16.5m. narrow are 15m and 7.5m .

**numbers below updated to match more precise field measurements**

Try creating rectangular grid and then use blanks to narrow it.

### Create grid

```{r}
stagger <- TRUE
buffer.beds <- 0
buffer.rows <- 2 # on each edge of each bed
exp.beds <- 11
beds <- letters[1:(buffer.beds+exp.beds)]
columns <- 4 # per bed
rows <- 63 # 9 of which are buffers
plants <- length(beds) * columns * rows
bedwidth <- 90 # cm
furrow <- 100 #cm
size <- 25 # plant diameter
radius <- size/2 

plan3 <- expand_grid(bed=beds,
                     column=1:columns,
                     row=1:rows,
                     radius=radius)

plan3 <- plan3 %>%
  mutate(plantedcol=(match(bed, letters)-1)*4 + column,
         bedcol = floor((match(bed, letters)-1) + 1   )) 

plan3
```

### add positions

```{r}
if (stagger) {
  column_offset <- sqrt((2*radius)^2 - radius^2) # Pythagorean theorem for offset spacing
  plan3 <- plan3 %>%
    mutate(y_pos=ifelse(is.even(column),
                        row*size, 
                        row*size-radius),
           x_pos=ifelse(column==1,
                        radius+(bedcol-1)*(bedwidth+furrow),
                        radius+(column-1)*column_offset+(bedcol-1)*(bedwidth+furrow)))
} else {
  plan3 <- plan3 %>%
    mutate(y_pos=row*size,
           x_pos=bedcol*(bedwidth+furrow))
}
plan3
```

### add indicator for whether or not a position is available and inbounds

```{r}
bedrows   <- c(63, 61, 60, 59, 53, 47, 45, 34, 24, 20, 16)

plan3 <- plan3 %>%
  mutate(status=ifelse(row <= bedrows[bedcol], "available", "out-of-bounds"))

# eliminate some beds for 2026 because we have fewer plants

plan3 <- plan3 %>%
  mutate(status=ifelse(str_detect(bed, "[abhijk]"), "out-of-bounds", status))

# block out tree position (approx)
plan3 <- plan3 %>%
  mutate(status=ifelse(bedcol == 5 & row > bedrows[5] -23 & row <= bedrows[5]-13, "out-of-bounds", status),
         status=ifelse(bedcol == 6 & row > bedrows[6] -19 & row <= bedrows[6]-9, "out-of-bounds", status)
         )

# block out positions with rocks

rocks <- tribble(~bed, ~column, ~row,
                 "a", 1, 44,
                 "a", 2, 44,
                 "a", 3, 45,
                 "a", 3, 45,
                 "a", 2, 48,
                 "a", 4, 48,
                 "b", 1, 35,
                 "c", 4, 3,
                 "c", 4, 4,
                 "c", 4, 5, #new rock/unplantable spot
                 "c", 1, 9,
                 "c", 4, 9, #new rock/unplantable spot
                 "c", 4, 10, #new rock/unplantable spot
                 "c", 3, 26,
                 "c", 4, 26,
                 "c", 3, 27,
                 "c", 4, 27,
                 "c", 1, 33,
                 "c", 3, 45,
                 "c", 3, 46,
                 #"d", 2, 5, no rock
                 "d", 1, 6,
                 "d", 3, 6,
                 "d", 4, 6,
                 "d", 1, 7,
                 "d", 2, 7,
                 "d", 4, 7,
                 "d", 2, 8,
                 "d", 1, 12,
                 "d", 2, 34,
                 "d", 3, 37,
                 "e", 3, 5, #new rock/unplantable spot
                 "e", 1, 6,
                 "e", 4, 6, #new rock/unplantable spot
                 #"e", 2, 9, no rock
                 "e", 2, 12, #new rock/unplantable spot
                 "e", 3, 12, #new rock/unplantable spot
                 "e", 2, 13,
                 "e", 3, 13, #new rock/unplantable spot
                 "e", 4, 15,
                 "e", 4, 16,
                 "e", 2, 26, #new rock/unplantable spot
                 "e", 1, 30,
                 "e", 2, 30,
                 "e", 3, 30,
                 #"e", 4, 30, no rock
                 "e", 1, 44, #new rock/unplantable spot
                 "e", 1, 47, #new rock/unplantable spot
                 "f", 2, 5,
                 "f", 3, 6,
                 #"f", 4, 17, no rock
                 "f", 4, 18,
                 "f", 4, 21, #new rock/unplantable spot
                 "f", 2, 40,
                 "g", 3, 15, #new rock/unplantable spot
                 "g", 4, 17, #new rock/unplantable spot
                 "g", 4, 22,
                 "g", 3, 23,
                 "g", 4, 23,
                 "g", 4, 29, #new rock/unplantable spot
                 "i", 1, 3,
                 "i", 3, 4,
                 "i", 4, 5,
                 "k", 3, 5,
                 )

rocks <- rocks %>% mutate(new_status="out-of-bounds") %>% arrange(bed, row, column)

plan3 <- left_join(plan3, rocks) %>% 
  mutate(status = ifelse(is.na(new_status), status, new_status)) %>%
  select(-new_status)

plan3
```

### Get 2023, 2024, and 2025 plants that were still alive in Oct2025

```{r, warning=FALSE}
survey <- read_csv("../input/WL2_2026_WinterSurv.csv") %>% 
  filter(Unique.ID!="buffer") %>% #remove buffers
  filter(is.na(death.date)) %>% #keep only plants without a death date
  select(bed:Unique.ID, death.date)
#120 alive post winter 

popinfo_2025 <- read_csv("../input/2025_Pop_Loc_Info Updated.csv") %>% 
  select(bed.block.order:bed, pop.id:mf, rep, Unique.ID) 

survey_pops <- survey %>% 
  left_join(popinfo_2025)

names(survey_pops)
```

```{r}
survey_pops_merge <- survey_pops %>%
  select(bed.block.order:column.order, bed, column=col, row, block_2025=block, status, pop.id:rep, Unique.ID)  %>%
  mutate(bed = tolower(bed),
         new_status=if_else(status=="available", "2025-survivor", status),
         column=case_match(column,
                        "A" ~ 1,
                        "B" ~ 2,
                        "C" ~ 3,
                        "D" ~ 4
                        )) %>% 
  select(-status) %>% 
  mutate(row=as.numeric(row))

plan3_oldplants <- left_join(plan3, survey_pops_merge) %>%
  mutate(status = ifelse(is.na(new_status), status, new_status)) %>%
  select(-new_status)
```

### Add buffers at the beginning and end of each bed

```{r}
plan3_buffers <- plan3_oldplants %>%
  
  #top buffers:
  mutate(buffer=(row<=5 & str_detect(bed, "[c-g]")),

  #bottom buffers
         buffer=ifelse(row > (bedrows[bedcol]-2) & row <= bedrows[bedcol] & str_detect(bed, "[c-g]"), TRUE, buffer),
         status=ifelse(buffer, "buffer", status)
  )
```

### CHECK THAT NO ALIVE PLANTS IN THE NEW BUFFER REGION
```{r}
plan3_buffers %>% filter(isTRUE(buffer)) %>% filter(!is.na(Unique.ID)) #none
plan3_buffers %>% filter(!is.na(Unique.ID), status=="available")
plan3_buffers %>% filter(!is.na(Unique.ID)) %>% 
  filter(status!="2025-survivor", status!="2024-survivor") #some 2025 plants in "buffer" spots
plan3_buffers <- plan3_buffers %>% 
  mutate(status=if_else(status=="buffer" & !is.na(Unique.ID), "2025-survivor", status))
```

### Plot it

```{r, fig.width=12, eval=TRUE}
plan3_buffers %>%
  ggplot(aes(x0=x_pos, y0=y_pos, r=radius, fill=status)) +
  geom_circle(alpha=.75) + 
  scale_fill_manual(values=c("out-of-bounds"="grey40", "available"="lightgreen", "2024-survivor"="blue", "2025-survivor"="magenta", "buffer"="yellow")) +
  coord_equal() + 
  #  geom_text(aes(x=x_pos, y=y_pos, label=bed), size= 14, data = bed.labels, inherit.aes = FALSE) + 
  theme(axis.text = element_text(size=18), axis.title = element_text(size=18)) +
  xlab("position (cm)") + ylab("position (cm)") +
  scale_y_reverse() +
  ggtitle(str_c("Total available positions = ", sum(plan3_buffers$status=="available")))

ggsave("../output/2026-WL2-FieldPlan.jpeg", width = 10, height = 6)
```

Position blocks in the available spots and initiate snaking 

```{r}
blocksizes <- germinated_blocks %>%
  group_by(block) %>%
  summarise(n=n()) %>%
  mutate(n=n+4) # a buffer row between blocks

blocks <-rep(blocksizes$block, time=blocksizes$n)

plan3_oddbeds <- plan3_buffers %>% filter(bed=="a" |bed=="c" | bed=="e" | bed=="g" | 
                                    bed=="i" | bed=="k") %>% 
  filter(status=="available") %>%
  arrange(bed, row, column) 
plan3_evenbeds <- plan3_buffers %>% filter(bed=="b" |bed=="d" | bed=="f" | bed=="h" | 
                                    bed=="j") %>% 
  filter(status=="available") %>%
  arrange(bed, desc(row), column) #%>% arrange(desc(row)) %>% arrange(column) #flip the even beds to allow snaking 

avail <- bind_rows(plan3_oddbeds, plan3_evenbeds) %>% 
  arrange(bed) %>% 
  magrittr::extract(1:length(blocks),) %>%
  mutate(block=blocks) %>%
  group_by(block) %>%
  mutate(block.id=str_c(block, str_pad(1:n(), width = 2, pad = "0"))) %>% 
  ungroup() %>% 
  select(bed:status, buffer:block.id)
```

## Assign Plants

### germinated

```{r}
set.seed(999)

germinated_blockpos <- germinated_blocks %>% 
  group_by(block) %>%
  slice_sample(n=nrow(.)) %>% # randomize position within blocks
  mutate(block.id=str_c(block, str_pad(1:n(), width = 2, pad = "0"))) %>%
  mutate(Unique.ID=as.character(Unique.ID)) %>% 
  ungroup() %>%
  arrange(block.id) %>%
  select(-reps)

germinated_avail <- left_join(avail, germinated_blockpos)

germinated_avail  %>% arrange(block, row, column) %>%
  select(block, row, column, block.id, everything())

germinated_avail  %>% arrange(pop.id) %>%# confirm different positions per block
  select(block, bed, row, column, pop.id)
germinated_avail %>% filter(block=="E") %>% arrange(bed)%>%
  select(block, bed, row, column, Unique.ID) #looks like snaking worked?
```

## combine bed map and field map

first get block rows correct and write to gsheets

```{r}
germinated_blockrow <- germinated_avail %>%
  ungroup() %>%
  arrange(block, row, column) %>%
  mutate(blockrow=row, blockcolumn=column) %>% 
  rename(bedrow=row) %>%
  group_by(block) %>%
  mutate(blockrow=as.numeric(as.factor(blockrow))) %>%
  ungroup()

germinated_blockrow %>% arrange(Rack, Unique.ID) %>%
  filter(!is.na(Unique.ID)) %>% 
  select(Rack, Unique.ID, pop.id, Rep, block, blockrow, blockcolumn) %>%
  write_sheet(popsheet,
              sheet="Randomization Map")
```


```{r}
plan3_oldplantsonly <- plan3_buffers %>%
  ungroup() %>%
  arrange(bedcol, row, column) %>%
  rename(bedrow=row) %>% 
  filter(status!="available") %>% 
  rename(Rep=rep) %>% 
  mutate(Rep=as.character(Rep)) %>% 
  select(bed:status, block_2025:buffer)

germinated_map <- germinated_blockrow %>% 
  select(bed, column, bedrow, radius, plantedcol, bedcol:block, Unique.ID, Type:Rep)
  

#combine
plan3_allplants <- bind_rows(plan3_oldplantsonly, germinated_map)
plan3_allplants %>% filter(status=="2025-survivor")
plan3_allplants %>% filter(status=="available")

# set up labels
plan3_plants_labels <- plan3_allplants %>%
  mutate(column=LETTERS[column]) %>%
  mutate(label=ifelse(buffer, "buff", Unique.ID),
    label=ifelse(status=="available" & pop.id=="buffer", "buff", label),
    label=ifelse(is.na(label), "", label),
    colorblock=ifelse(is.na(block), status, block)) %>% 
  mutate(bed=toupper(bed))


#Write it out 

plan3_plants_labels %>% arrange(pop.id, Rep) %>%
  select(pop.id, Rep, everything() ) %>%
  write_sheet(popsheet,
              sheet="Buffer New Bed Map")
```

```{r, fig.height=12}
pdf("../output/2026-WL2-planting_maps.pdf",width = 8.5, height = 11)

colors=c("grey40", "yellow", "blue", "magenta", rep(c("orange", "skyblue", "green"), length.out=length(germblocks))) %>%
  set_names("out-of-bounds", "buffer", "2024-survivor", "2025-survivor", germblocks)


for(b in unique(plan3_plants_labels$bed)) {
  pl1 <- plan3_plants_labels %>% filter(bed==b, bedrow<33) %>%
    ggplot(aes(x0=x_pos, y0=y_pos, r=radius, label=label, fill=colorblock)) +
    scale_fill_manual(values=colors) +
    geom_circle(alpha=.5, color="grey70") +
    geom_text(aes(x=x_pos, y=y_pos), size=3, lineheight=.8) +
    theme(axis.text.x = element_text(angle=90, size=12, vjust = 0.5)) +
    xlab("position (cm)") + ylab("position (cm)") +
    scale_y_reverse() +
    coord_equal(ratio=.4) + 
    ggtitle(str_c("Bed ", b, ", sheet 1"))
  print(pl1)
  pl2 <- plan3_plants_labels %>% filter(bed==b, bedrow>=33) %>%
    ggplot(aes(x0=x_pos, y0=y_pos, r=radius, label=label, fill=colorblock)) +
    scale_fill_manual(values=colors) +
    geom_circle(alpha=.5, color="grey70") +
    geom_text(aes(x=x_pos, y=y_pos), size=3, lineheight=.8) +
    theme(axis.text.x = element_text(angle=90, size=12, vjust = 0.5)) +
    xlab("position (cm)") + ylab("position (cm)") +
    scale_y_reverse() +
    coord_equal(ratio=.4) + 
    ggtitle(str_c("Bed ", b, ", sheet 2"))
  print(pl2)
}

dev.off()
```

Randomization maps

```{r, fig.height=11, fig.width=8.5}

pdf("../output/2026-WL2-randomization_maps.pdf",width = 8.5, height = 11)

germinated_random_map <- germinated_blockrow %>% arrange(block, blockrow, blockcolumn)

for(b in unique(germinated_random_map$block)) {
  pl <- germinated_random_map %>% filter(block==b) %>%
    mutate(label=ifelse(pop.id=="buffer", "buff", Unique.ID)) %>%
    mutate(label=ifelse(is.na(label), "buff", label)) %>%
    ggplot() +
    geom_label(aes(x=blockcolumn, y=blockrow, label=label), size=3, lineheight=.8) +
    theme(axis.text = element_text(size=12)) +
    scale_y_reverse(breaks = 1:31) +
    scale_x_discrete(position = "top") +
    ggtitle(str_c("Block ", b)) +
    coord_equal(ratio=.5)
  print(pl)
}

dev.off()

```
