---
title: "Common Garden Design"
output: html_notebook
---

## Intro

Goal: decide number of pops, maternal families, etc.  Then planting design.
Parameters: maximize number of pops, then number of maternal families.  Plant 2000.

```{r}
library(tidyverse)
library(googlesheets4)
library(ggforce)

is.even <- function(x) x%%2 == 0
```

## get data on pops and families

```{r}
pops <- read_sheet("https://docs.google.com/spreadsheets/d/1dif9Y5hbkSa56Bgonj04-jXh8jNc6f13RBS6BPUf1IQ",
                   skip=1,
                   na=c("NA", ""),
                   col_types = c("ciiiccccc"),
                   .name_repair = "universal") %>%
  mutate(approx.number.seeds = as.integer(str_remove_all(approx.number.seeds,"[^0-9]")))

pops
```
## filter to one entry per pop, etc

```{r}
pops.filtered <- pops %>% group_by(parent.pop) %>% slice_max(order_by=maternal.families) %>%
  filter(approx.number.seeds >= 100) %>%
  filter(!(parent.pop %in% c("HH", "RB"))) # old seed

pops.filtered %>% arrange(maternal.families)
```

```{r}
sum(pops.filtered$maternal.families>=8)
```

```{r}
sum(pops.filtered$maternal.families>=15)
```

### Scenario 1:

Plant 21 pops * 8 families * 12 reps (= 2016 plants)

### Scenario 2:

Plant 11 pops * 15 families * 12 reps (= 1980 plants)

### Scenario 3:

__Note: after calculating above, Sarah found more mfs from WV, so:_

3 mfs from WR and 7 mfs from everyone else

total mfs = 3+7*22 = 157 * 13 reps (= 2041 plants)

## Which pops?

The group decision was to go with scenario 3: "3 mfs from WV, 4 mfs WR and 7 mfs from everyone else"

This is the full (filtered) data sheet.

```{r}
pops.filtered %>% arrange(parent.pop) %>% write_csv("../output/Pops_for_2022_UCD.csv")
```

## planting grid

### go with plan 3

### Example plot for "60 in veg crop" bed.

Sarah says this is 100 cm of plantable width.  

2041 plants.  1 block per rep, so 13 blocks each of 157 plants

157/4 = 39.25 with 4 columns this is 39.25 plants per column per block.  Call it 40.  But then two more buffers on each side, call it 44

Create grid
```{r}
stagger <- TRUE
blocks <- 15 # two of which are buffers
columns <- 4 # per block
rows <- 44 # four of which are buffers
plants <- blocks * columns * rows
bedwidth <- 100 # cm
furrow <- 52 #cm
size <- 30 # plant diameter
radius <- size/2 
blankevery <- 2 # what is the spacing between blank beds?  (2 means every other; 3 means every third)
expandfactor <- blankevery/(blankevery-1)

plan3 <- expand_grid(block=LETTERS[1:blocks],
                     column=1:columns,
                     row=1:rows,
                     radius=radius)

plan3 <- plan3 %>%
  mutate(plantedcol=(match(block, LETTERS)-1)*4 + column,
         bedcol = floor((match(block, LETTERS)-1)*expandfactor + 1   )) 

plan3
```

add positions
```{r}
if (stagger) {
  ## first part needs to be updated for beds
  column_offset <- sqrt((2*radius)^2 - radius^2) # Pythagorean theorem for offset spacing
  plan3 <- plan3 %>%
    mutate(y_pos=ifelse(is.even(column),
                        row*size, 
                        row*size-radius),
           x_pos=ifelse(column==1,
                        radius+(bedcol-1)*(bedwidth+furrow),
                        radius+(column-1)*column_offset+(bedcol-1)*(bedwidth+furrow)))
} else {
  plan3 <- plan3 %>%
    mutate(y_pos=row*size,
           x_pos=bedcol*(bedwidth+furrow))
}
plan3
```
Plot it
```{r, fig.width=12}

#create block labels
# block.labels <- plan3 %>%
#   group_by(block) %>%
#   summarize(x_pos = mean(x_pos+block_x_offset), y_pos=max(y_pos*1.075)) 


plan3 %>% #filter(block=="A", row <6) %>%
  ggplot(aes(x0=x_pos, y0=y_pos, r=radius)) +
  geom_circle(fill="lightgreen", alpha=.25) + 
  coord_equal() + 
#  geom_text(aes(x=x_pos, y=y_pos, label=block), size= 14, data = block.labels, inherit.aes = FALSE) + 
  theme(axis.text = element_text(size=18), axis.title = element_text(size=18)) +
  xlab("position (cm)") + ylab("position (cm)")

# ggsave("../output/FieldPlan.jpeg", width = 10, height = 6)
```

Plot it in feet
```{r, fig.width=12}

#create block labels
# block.labels <- plan3 %>%
#   group_by(block) %>%
#   summarize(x_pos = mean(x_pos+block_x_offset), y_pos=max(y_pos*1.075)) 


plan3 %>% #filter(block=="A", row <6) %>%
  mutate(across(.cols=c(x_pos, y_pos, radius), ~ .x*0.0328084)) %>% # convert cm to feet
  ggplot(aes(x0=x_pos, y0=y_pos, r=radius)) +
  geom_circle(fill="lightgreen", alpha=.25) + 
  coord_equal(xlim = c(4,145)) + 
#  geom_text(aes(x=x_pos, y=y_pos, label=block), size= 14, data = block.labels, inherit.aes = FALSE) + 
  theme(axis.text = element_text(size=18), axis.title = element_text(size=18)) +
  xlab("position (feet)") + ylab("position (feet)") 

# ggsave("../output/FieldPlan.jpeg", width = 10, height = 6)
```


Alternate, what if we stack blocks

```{r, fig.width=12}
# only offset x_positions (1 "row" of blocks)
plan3 <- plan3 %>%
  mutate(block_x_offset = as.integer(as.factor(block)) %% 7 -1,
         block_x_offset = block_x_offset* (aisle + size + (columns-1)*column_offset)) %>%
  mutate(block_y_offset = (as.integer(as.factor(block)) %/% 7)*-1+1 ,
         block_y_offset = block_y_offset*size*rows+block_y_offset*aisle*1.5)

#create block labels
block.labels <- plan3 %>%
  group_by(block) %>%
  summarize(x_pos = mean(x_pos+block_x_offset), y_pos=max(block_y_offset + y_pos*1.075))


plan3 %>% #filter(block=="A", row <6) %>%
  ggplot(aes(x0=x_pos+block_x_offset, y0=y_pos+block_y_offset, r=radius)) +
  geom_circle(fill="lightgreen", alpha=.25) + 
  coord_equal() + 
  geom_text(aes(x=x_pos, y=y_pos, label=block), size= 20, data = block.labels, inherit.aes = FALSE) + 
  theme(axis.text = element_text(size=16), axis.title = element_text(size=16)) +
  xlab("position (cm)") + ylab("position (cm)")
```

```{r}
plants  <- read_sheet("https://docs.google.com/spreadsheets/d/1EaSA2QrlWAjFR35pedcvheQp3X7Agsla6pZtE2C1QJ4",
                      sheet="label making")

plants 
```
set up row labels for sampling 
```{r}
rows.sample <- rep(1:rows, times=c(rep(columns,rows-1), 1) )

#cols.sample <- rep(LETTERS[1:columns], times=c(rows, rep(rows-1, columns-1)))
```

Add block (easy because block = rep, but I am using letters for block)
```{r}
plants <- plants %>% mutate(block=LETTERS[rep])
```
now add row and column positions:

```{r}
set.seed(092222) # so we get the same "randomization" each time we run this
plants <- plants %>%
  group_by(block) %>%
  mutate(row=sample(rows.sample)) %>%
  group_by(block, row) %>%
  mutate(col=sample(LETTERS[1:columns], size=n()))

plants %>% arrange(parent.pop, mf) # confirm different positions per block
```

now add flat into for the growth chamber.  flats are 7 X 14.  We don't have space to do 4 rows per flat, so we will do 7 rows per flat and spread them out day of planting

We have a total of 13 * 4 = 52 columns.  8 flats have a total of 56 columns.

```{r}
plants <- plants %>% ungroup() %>%
  mutate(ubercolumn = as.integer(factor(str_c(block, col))),
         flat = (ubercolumn + 6) %/% 7, # column assignments to flat (7 columns per flat)
         flat = flat + ((row-1) %/% 14)*8) # adjust for row (14 rows per flat, 8 flats per row of flats)

plants

tail(plants)
```

plot flatmaps
```{r, fig.height=10, fig.width=7.5}

flat.plots <- map(unique(plants$flat), function(f) {
  
  plants.flat <- plants %>% filter(flat==f) %>%
    mutate(label=str_c(parent.pop, "-", mf, "\n", block, "-", row, col))
  
  row.labels <- plants.flat %>% pull(row) %>% unique() %>% sort()
  col.labels <- plants.flat %>% pull(ubercolumn) %>% unique() %>% sort()
  
  plants.flat %>%  ggplot(aes(x=ubercolumn, y=row, x0=ubercolumn, y0 = row, label=label)) +
    geom_tile(fill="white", color="black") +
    coord_equal(expand = FALSE) +
    geom_text(size=3) +
    scale_y_reverse(labels=row.labels, breaks=row.labels) +
    scale_x_continuous(labels=col.labels, breaks=col.labels) +
    ggtitle(str_c("Flat ", f)) 
}
)
```

```{r, fig.height=10, fig.width=7.5}
flat.plots[[1]]
```

```{r}
pdf(file="../output/flatmaps.pdf", width=7.5, height = 10)
walk(flat.plots, print)
dev.off()
```

```{r}
plants %>% arrange(parent.pop, mf, rep) %>% write_sheet("https://docs.google.com/spreadsheets/d/1EaSA2QrlWAjFR35pedcvheQp3X7Agsla6pZtE2C1QJ4",
                                                        sheet="Flat and Block map")
```

