---
title: "Common Garden Design"
output: html_notebook
---

## Intro

Determine randomization strategy and plant placing when we have had uneven germination of our original plan.

This is written for UCD Fall 2022 planting.

```{r}
library(tidyverse)
library(googlesheets4)
library(ggforce)

is.even <- function(x) x%%2 == 0
```

## get data on germinated seed

```{r}
# the original reps
seed <- read_sheet("https://docs.google.com/spreadsheets/d/1EaSA2QrlWAjFR35pedcvheQp3X7Agsla6pZtE2C1QJ4/edit#gid=1798047255", 
                   sheet = "germination",
                   na=c("NA", "")
) %>% 
  select(parent.pop, mf, rep, `germinated?`)
seed
```


```{r}
# extra seed
extra <- read_sheet("https://docs.google.com/spreadsheets/d/1EaSA2QrlWAjFR35pedcvheQp3X7Agsla6pZtE2C1QJ4/edit#gid=1798047255", 
                   sheet = "extra mf reps",
                   na=c("NA", "")
) %>%
  filter(`germinated?` == 1) # we are not going to use ungerminated extra seed.
extra
```

```{r}
seed.combined <- bind_rows(seed, extra)

germinated <- seed.combined %>%
  filter(`germinated?`==1)

not.germinated <- seed.combined %>%
  filter(`germinated?` == 0)
```

## assign blocks

Can plant 160  plants per bed
```{r}
nrow(germinated) 
nrow(germinated) / 160
```

get a histogram of rep count

```{r}
germinated %>%
  group_by(parent.pop, mf) %>%
  summarize(n=n()) %>%
  ggplot(aes(x=n)) +
  geom_histogram(binwidth = 1)
```

```{r}
germinated %>%
  group_by(parent.pop) %>%
  summarize(n=n()) %>%
  arrange(n)
```

A bit challenging because for pops with few reps total we should be careful to distribute the pops across the blocks.

I think one way this can work is to arrange by pops and mfs and then assign a randomized set of block labels to it.

```{r}
germbeds <- c("C", "E", "G", "I", "K")
notgermbeds <- LETTERS[1:13] %>% setdiff(germbeds)

germblocks <- str_c(rep(germbeds,2), c("1", "2")) %>% sort

germreps <- ceiling(nrow(germinated) / 10 ) # number of times we have to assign blocks

# set.seed(777)
# germinated <- germinated %>%
#   arrange(parent.pop, mf) %>%
#   mutate(block= {map(1:germreps, ~ sample(germblocks)) %>%
#       unlist() %>% magrittr::extract(1:nrow(germinated))}, # clean up last rep
#       bed=str_sub(block, 1,1)
#       )


germinated$block <- rep(germblocks, length.out=nrow(germinated))
#for low rep mfs, randomize the whole pop
set.seed(777)
germinated_lowrep <- germinated %>%
  group_by(parent.pop, mf) %>%
  mutate(reps=n()) %>%
  filter(reps < 10) %>%
  group_by(parent.pop) %>%
  mutate(block= {map(1:ceiling(n()/10), ~ sample(germblocks)) %>%
       unlist() %>% magrittr::extract(1:n())}, # clean up last rep
         bed=str_sub(block, 1,1)) %>%
  ungroup() 

#For high rep mfs, just use the sequentially assigned blocks
germinated_highrep <- germinated %>%
  group_by(parent.pop, mf) %>%
  mutate(reps=n()) %>%
  filter(reps >= 10) %>%
  ungroup %>%
  mutate(bed=str_sub(block, 1,1))

germinated <- bind_rows(germinated_lowrep, germinated_highrep) %>% arrange(parent.pop, mf)  


# set.seed(777)
# germinated <- germinated %>%
#   arrange(parent.pop, mf) %>%
#   mutate(block= rep(germblocks, length.out = 757)) %>%
#   group_by(parent.pop) %>%
#   mutate(block=sample(block), #sampling within pop
#          bed=str_sub(block, 1,1)) %>%
#   ungroup()


germinated  
```
reality check distribution across mfs
```{r}
germinated %>%
  pivot_wider(id_cols = c(parent.pop, mf), names_from = block, values_from = `germinated?`, values_fill = 0, values_fn = sum)
```

reality check distribution across parent.pops
```{r}
germinated %>%
  pivot_wider(id_cols = parent.pop, names_from = block, values_from = `germinated?`, values_fill = 0, values_fn = sum)
```

plants per block:
```{r}
germinated %>%
  group_by(block) %>%
  summarise(n=n())
```


## filter to one entry per pop, etc

```{r}
pops.filtered <- pops %>% group_by(parent.pop) %>% slice_max(order_by=maternal.families) %>%
  filter(approx.number.seeds >= 100) %>%
  filter(!(parent.pop %in% c("HH", "RB"))) # old seed

pops.filtered %>% arrange(maternal.families)
```

```{r}
sum(pops.filtered$maternal.families>=8)
```

```{r}
sum(pops.filtered$maternal.families>=15)
```

### Scenario 1:

Plant 21 pops * 8 families * 12 reps (= 2016 plants)

### Scenario 2:

Plant 11 pops * 15 families * 12 reps (= 1980 plants)

### Scenario 3:

__Note: after calculating above, Sarah found more mfs from WV, so:_

3 mfs from WR and 7 mfs from everyone else

total mfs = 3+7*22 = 157 * 13 reps (= 2041 plants)

## Which pops?

The group decision was to go with scenario 3: "3 mfs from WV, 4 mfs WR and 7 mfs from everyone else"

This is the full (filtered) data sheet.

```{r}
pops.filtered %>% arrange(parent.pop) %>% write_csv("../output/Pops_for_2022_UCD.csv")
```

## planting grid

### go with plan 3

### Example plot for "60 in veg crop" bed.

Sarah says this is 100 cm of plantable width.  

2041 plants.  1 block per rep, so 13 blocks each of 157 plants

157/4 = 39.25 with 4 columns this is 39.25 plants per column per block.  Call it 40.  But then two more buffers on each side, call it 44

Create grid
```{r}
stagger <- TRUE
buffer.blocks <- 2
buffer.rows <- 2 # on each edge of each block
exp.blocks <- 13
blocks <- LETTERS[1:(buffer.blocks+exp.blocks)]
columns <- 4 # per block
rows <- 44 # four of which are buffers
plants <- length(blocks) * columns * rows
bedwidth <- 100 # cm
furrow <- 52 #cm
size <- 30 # plant diameter
radius <- size/2 
blankevery <- 2 # what is the spacing between blank beds?  (2 means every other; 3 means every third)
expandfactor <- blankevery/(blankevery-1)

plan3 <- expand_grid(block=blocks,
                     column=1:columns,
                     row=1:rows,
                     radius=radius)

plan3 <- plan3 %>%
  mutate(plantedcol=(match(block, LETTERS)-1)*4 + column,
         bedcol = floor((match(block, LETTERS)-1)*expandfactor + 1   )) 

plan3
```

add positions
```{r}
if (stagger) {
  ## first part needs to be updated for beds
  column_offset <- sqrt((2*radius)^2 - radius^2) # Pythagorean theorem for offset spacing
  plan3 <- plan3 %>%
    mutate(y_pos=ifelse(is.even(column),
                        row*size, 
                        row*size-radius),
           x_pos=ifelse(column==1,
                        radius+(bedcol-1)*(bedwidth+furrow),
                        radius+(column-1)*column_offset+(bedcol-1)*(bedwidth+furrow)))
} else {
  plan3 <- plan3 %>%
    mutate(y_pos=row*size,
           x_pos=bedcol*(bedwidth+furrow))
}
plan3
```
Plot it
```{r, fig.width=12}

#create block labels
# block.labels <- plan3 %>%
#   group_by(block) %>%
#   summarize(x_pos = mean(x_pos+block_x_offset), y_pos=max(y_pos*1.075)) 


plan3 %>% #filter(block=="A", row <6) %>%
  ggplot(aes(x0=x_pos, y0=y_pos, r=radius)) +
  geom_circle(fill="lightgreen", alpha=.25) + 
  coord_equal() + 
  #  geom_text(aes(x=x_pos, y=y_pos, label=block), size= 14, data = block.labels, inherit.aes = FALSE) + 
  theme(axis.text = element_text(size=18), axis.title = element_text(size=18)) +
  xlab("position (cm)") + ylab("position (cm)")

ggsave("../output/FieldPlan.jpeg", width = 10, height = 6)
```

Plot it in feet
```{r, fig.width=12}

#create block labels
# block.labels <- plan3 %>%
#   group_by(block) %>%
#   summarize(x_pos = mean(x_pos+block_x_offset), y_pos=max(y_pos*1.075)) 


plan3 %>% #filter(block=="A", row <6) %>%
  mutate(across(.cols=c(x_pos, y_pos, radius), ~ .x*0.0328084)) %>% # convert cm to feet
  ggplot(aes(x0=x_pos, y0=y_pos, r=radius)) +
  geom_circle(fill="lightgreen", alpha=.25) + 
  coord_equal(xlim = c(4,145)) + 
  #  geom_text(aes(x=x_pos, y=y_pos, label=block), size= 14, data = block.labels, inherit.aes = FALSE) + 
  theme(axis.text = element_text(size=18), axis.title = element_text(size=18)) +
  xlab("position (feet)") + ylab("position (feet)") 

# ggsave("../output/FieldPlan.jpeg", width = 10, height = 6)
```
Output spacing for Black A to use as guide for cuting holes
```{r, fig.width=8, fig.height=10.5}

x_labels <- plan3 %>% filter(block=="A") %>% pull(x_pos) %>% round(1) %>% unique()

plan3 %>% filter(block=="A") %>%
  ggplot(aes(x0=x_pos, y0=y_pos, r=radius, label=y_pos)) +
  geom_circle(fill="lightgreen", alpha=.25, color="grey70") + 
  geom_text(aes(x=x_pos, y=y_pos), size=3) +
  scale_x_continuous(breaks = x_labels) +
  coord_equal(ylim = c(20,660)) + 
  #  geom_text(aes(x=x_pos, y=y_pos, label=block), size= 14, data = block.labels, inherit.aes = FALSE) + 
  theme(axis.text.x = element_text(angle=90, size=12, vjust = 0.5)) +
  xlab("position (cm)") + ylab("position (cm)") 
ggsave("../output/spacing_guide_1.pdf", width = 8, height = 10.5)

plan3 %>% filter(block=="A") %>%
  ggplot(aes(x0=x_pos, y0=y_pos, r=radius, label=y_pos)) +
  geom_circle(fill="lightgreen", alpha=.25, color="grey70") + 
  geom_text(aes(x=x_pos, y=y_pos), size=3) +
  scale_x_continuous(breaks = x_labels) +
  coord_equal(ylim = c(700,1320)) + 
  #  geom_text(aes(x=x_pos, y=y_pos, label=block), size= 14, data = block.labels, inherit.aes = FALSE) + 
  theme(axis.text.x = element_text(angle=90, size=12, vjust = 0.5)) +
  xlab("position (cm)") + ylab("position (cm)") 
ggsave("../output/spacing_guide_2.pdf", width = 8, height = 10.5)
```


Alternate, what if we stack blocks (NOT UPDATED)

```{r, fig.width=12}
# # only offset x_positions (1 "row" of blocks)
# plan3 <- plan3 %>%
#   mutate(block_x_offset = as.integer(as.factor(block)) %% 7 -1,
#          block_x_offset = block_x_offset* (aisle + size + (columns-1)*column_offset)) %>%
#   mutate(block_y_offset = (as.integer(as.factor(block)) %/% 7)*-1+1 ,
#          block_y_offset = block_y_offset*size*rows+block_y_offset*aisle*1.5)
# 
# #create block labels
# block.labels <- plan3 %>%
#   group_by(block) %>%
#   summarize(x_pos = mean(x_pos+block_x_offset), y_pos=max(block_y_offset + y_pos*1.075))
# 
# 
# plan3 %>% #filter(block=="A", row <6) %>%
#   ggplot(aes(x0=x_pos+block_x_offset, y0=y_pos+block_y_offset, r=radius)) +
#   geom_circle(fill="lightgreen", alpha=.25) + 
#   coord_equal() + 
#   geom_text(aes(x=x_pos, y=y_pos, label=block), size= 20, data = block.labels, inherit.aes = FALSE) + 
#   theme(axis.text = element_text(size=16), axis.title = element_text(size=16)) +
#   xlab("position (cm)") + ylab("position (cm)")
```
## Assign Plants
```{r}
plants  <- read_sheet("https://docs.google.com/spreadsheets/d/1EaSA2QrlWAjFR35pedcvheQp3X7Agsla6pZtE2C1QJ4",
                      sheet="label making")

plants 
```
set up row labels for sampling 
```{r}
# because we have two buffer rows at the beginning and end, we start at row 3 and go to rows -2
rows.sample <- rep(3:(rows-2), times=c(rep(columns,rows-5), 1) ) 

#cols.sample <- rep(LETTERS[1:columns], times=c(rows, rep(rows-1, columns-1)))
```

Add block (easy because block ~= rep, but I am using letters for block and need to adjust for the buffer rows)
```{r}
plants <- plants %>% mutate(block=LETTERS[rep+(buffer.blocks/2)])
```
now add row and column positions:

```{r}
set.seed(092222) # so we get the same "randomization" each time we run this
plants <- plants %>%
  group_by(block) %>%
  mutate(row=sample(rows.sample)) %>%
  group_by(block, row) %>%
  mutate(column=sample(LETTERS[1:columns], size=n()))

plants %>% arrange(parent.pop, mf) # confirm different positions per block
```
Fill in the blanks...
```{r}
plants <- plants %>% ungroup() %>%
  expand(block, row=1:rows, column) %>% 
  left_join(plants) %>%
  mutate(parent.pop=ifelse(is.na(parent.pop), "buffer", parent.pop))
plants
```


now add flat info for the growth chamber.  flats are 7 X 14.  We don't have space to do 4 rows per flat, so we will do 7 rows per flat and spread them out day of planting

We have a total of 13 * 4 = 52 columns.  8 flats have a total of 56 columns.

```{r, eval=TRUE}
plants <- plants %>% ungroup() %>%
  mutate(ubercolumn = as.integer(factor(str_c(block, column))),
         flat = ifelse(row <= rows - buffer.rows, ((row - 1 ) %/% 14) + 1, NA), # row assignment for lat
         flat = flat + ((ubercolumn - 1) %/% 7) * 3,  # column assignments to flat (7 columns per flat)
         flat_row = ((row-1) %% 14) + 1,
         flat_column = LETTERS[((ubercolumn-1) %% 7) + 1])
plants

tail(plants)
```

Not using.  Can use if we want flats to run the other way in the field.
```{r, eval=FALSE}
plants <- plants %>% ungroup() %>%
  mutate(ubercolumn = as.integer(factor(str_c(block, column))),
         flat = (ubercolumn + 6) %/% 7, # column assignments to flat (7 columns per flat)
         flat = flat + ((row-1) %/% 14)*8, # adjust for row (14 rows per flat, 8 flats per row of flats)
         flat_row = ((row-1) %% 14) + 1,
         flat_column = LETTERS[((ubercolumn-1) %% 7) + 1)]

plants

tail(plants)
```


plot flatmaps
```{r, fig.height=10, fig.width=7.5}

flat.plots <- map(unique(plants$flat), function(f) {
  
  plants.flat <- plants %>% filter(flat==f) %>%
    mutate(label=ifelse(is.na(mf),
                        str_c(parent.pop, "\n", block, "-", row, column),
                        str_c(parent.pop, "-", mf, "-r", rep, "\n", block, "-", row, column)))
  
  row.labels <- plants.flat %>% pull(row) %>% unique() %>% sort()
  col.labels <- plants.flat %>% pull(ubercolumn) %>% unique() %>% sort()
  
  plants.flat %>%  ggplot(aes(x=ubercolumn, y=row, x0=ubercolumn, y0 = row, label=label)) +
    geom_tile(fill="white", color="black") +
    coord_equal(expand = FALSE) +
    geom_text(size=3) +
    scale_y_reverse(labels=row.labels, breaks=row.labels) +
    scale_x_continuous(labels=col.labels, breaks=col.labels) +
    ggtitle(str_c("Flat ", f)) 
}
)
```

```{r, fig.height=10, fig.width=7.5}
flat.plots[[1]]
```

```{r}
pdf(file="../output/flatmaps.pdf", width=7.5, height = 10)
walk(flat.plots, print)
dev.off()
```

```{r}
plants %>% arrange(parent.pop, mf, rep) %>%
  select(parent.pop, mf, rep, flat, flat_row, flat_column, everything() ) %>% write_sheet("https://docs.google.com/spreadsheets/d/1EaSA2QrlWAjFR35pedcvheQp3X7Agsla6pZtE2C1QJ4",
                                                        sheet="Flat and Block map")
```

## combine flat map and field map

Plot with flats super-imposed
```{r, fig.width=12}

plan3_flats <- plan3 %>%
  mutate(column=LETTERS[column]) %>%
  full_join(plants)


flat_coords <- plan3_flats %>%
  filter(!is.na(flat)) %>%
  group_by(flat) %>%
  summarize(
    xmin=min(x_pos)-radius,
    xmax=max(x_pos)+radius,
    ymin=min(y_pos)-radius,
    ymax=max(y_pos)+radius
  )

field.flat.plot <- flat_coords %>%
  ggplot(aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=as.factor(flat %% 12))) +
  geom_rect() +
  scale_fill_brewer(palette = "Paired", guide="none") +
  geom_label(aes(x=xmin, y=ymin, label=flat), nudge_x = 150, nudge_y = -100, fill="grey90") +
  coord_equal() +
  scale_y_reverse()

field.flat.plot +  
  geom_circle(data = plan3_flats, aes(x0=x_pos, y0=y_pos, r=radius), fill="lightgreen", color="grey50", alpha=.25, inherit.aes = FALSE) + 
  #  geom_text(aes(x=x_pos, y=y_pos, label=block), size= 14, data = block.labels, inherit.aes = FALSE) + 
  theme(axis.text = element_text(size=18), axis.title = element_text(size=18)) +
  xlab("position (cm)") + ylab("position (cm)") 

 ggsave("../output/FieldPlan_with_flats.jpeg", width = 10, height = 6)
```

